### 2016-07-21 doc/main.tex
doc

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Application Note}

\usepackage{url}

\begin{document}
\firstpage{1}

\subtitle{Application Note}

\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: phylogenetic ancestral reconstruction}
\author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
\address{$^{\text{\sf 1}}$Department of Bioengineering, University of California, Berkeley, 94703, USA.}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX } %; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{
{\bf Motivation.}
{\bf Results.}
{\bf Availability and Implementation.}
Indel Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
It is written in C++11.
{\bf Contact.}
Ian Holmes {\tt ihholmes+indelhistorian@gmail.com}.
{\bf Supplementary Information.}
None.
}

\maketitle

\section{Introduction}

\begin{methods}
\section{Methods}

\cite{Westesson2012-zg}

\end{methods}

\section{Availability}

Indel Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license. It is written in C++11, and compiles on a POSIX system with Clang (v.6.1.0). It requires the GNU Scientific Software library (libgsl), the ZLib compression library (libz), and the Boost library.

\section*{Acknowledgements}

Indel Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort), and the StackOverflow community.

\section*{Funding}

This work has been supported by NHGRI grant R01-HG004483.

\bibliographystyle{natbib}
%\bibliographystyle{bioinformatics}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\bibliography{references}
\end{document}

### 2016-07-21 doc/main.tex
writeup

@@
 
+Multiple sequence alignments are used for different purposes in bioinformatics,
+and these different applications have led to different criteria for evaluating
+the accuracy of multiple alignment tools
+
+Homology modeling in protein structure prediction
+leads to accuracy metrics that quantify the number of correctly aligned residues
+compared to a benchmark dataset of structurally-informed reference alignments
+
+Significant progress has been made optimizing these metrics
+
+Successful techniques include using Bayesian decision theory to maximize the expected
+score under the metrics \cite{TCOFFEE,ProbCons,AMAP,FSA}
+and fine-tuning the alignment scoring function to reflect known details of
+protein sequences under selection for folding, such as the reduced likelihood of indels in
+hydrophobic regions \cite{MAFFT,MUSCLE}
+
+These have been accompanied by performance optimizations at the rate-limiting steps
+of multiple alignment, including all-versus-all sequence comparison \cite{FSA,MUSCLE}
+
+The study of sequence evolution and molecular phylogenetics gives rise to
+an alternative set of criteria for measuring alignment accuracy
+
+Previous work showed that, for the purposes of estimating
+molecular evolutionary parameters
+such as the rates of insertion-deletion events \cite{Westesson2012-zg}
+or the ratio between synonymous abd nonsynonymous substitution rates \cite{Redelings},
+it is best to use an explicit probabilistic model of the sequence evolution process
+and to treat sequence alignment rigorously as a missing data estimation problem
+under this model
+
+The ProtPal program is one such approach which uses the theory of
+weighted finite-state transducers
+These are automata which can be multiplied like substitution matrices
+and so can be thought of as a strict generalization of Felsenstein's likelihood-phylogenetic
+approach to multiple sequence alignment
+(in fact, the use of transducers makes Felsenstein's pruning algorithm formally
+equivalent to a special case of Sankoff's multiple alignment algorithm)
+
+A previously published benchmark of alignment methods on simulated sequence data
+showed that most alignment algorithms introduce systematic biases into the
+estimation of indel rates, and that these biases are typically substantially worse
+at higher indel rates.
+
+ProtPal was found to be the most accurate algorithm for indel rate estimation,
+followed by PRANK and then MUSCLE.
+
+However, the implementation of ProtPal published with that benchmark was too slow
+for practical use, and no benchmark of ProtPal on structural alignment benchmarks was available
+
+Here, we present a clean implementation of the algorithm underlying ProtPal
+in a new tool, Indel Historian, which we recommend for evolutionary applications such
+as ancestral sequence reconstruction and molecular phylogenetics
+
+We also report the results of benchmarking ProtPal on the BAliBase 3.0 benchmark
+
 \begin{methods}
@@
 
+Comparison to MCMC tools
+
+\section{Results}
+
 \section{Availability}
@@
 It is written in C++11, and compiles on a POSIX system with Clang (v.6.1.0).
-It requires the GNU Scientific Software library (libgsl), the ZLib compression library (libz), and the Boost library.
+It requires the GNU Scientific Software library (libgsl), the ZLib compression library (libz), and the Boost C++ library.
 

### 2016-07-21 doc/main.tex
writeup

@@
 Here, we present a clean implementation of the algorithm underlying ProtPal
-in a new tool, Indel Historian, which we recommend for evolutionary applications such
+in a new tool, Indel Historian
+The tool is targeted to evolutionary applications such
 as ancestral sequence reconstruction and molecular phylogenetics
@@
 Comparison to MCMC tools
+BaliPhy, PhylogenyCafe, HandAlign
 
 \section{Results}
+
+\section{Discussion}
+
+The best-performing method MUSCLE introduces heuristics into its scoring scheme
+that improve accuracy.
+
+It is harder to incorporate these sorts of modifications into a rigorous phylogenetic model
+primarily because they violate the assumption of independence
+between indel and substitution processes
+that most such models rely on
+
+Decision theory may also be useful in conjunction with MCMC tools \cite{HermanHeinEtc}
 

### 2016-07-22 doc/main.tex
refs

@@
 Successful techniques include using Bayesian decision theory to maximize the expected
-score under the metrics \cite{TCOFFEE,ProbCons,AMAP,FSA}
+score under the metrics \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
 and fine-tuning the alignment scoring function to reflect known details of
 protein sequences under selection for folding, such as the reduced likelihood of indels in
-hydrophobic regions \cite{MAFFT,MUSCLE}
+hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}
+
+PRANK\citep{LoytynojaGoldman2005}, Muscle \citep{Edgar2004b}, ClustalW\citep{LarkinEtAl2007}, MAFFT\citep{KatohEtAl2005}
 
 These have been accompanied by performance optimizations at the rate-limiting steps
-of multiple alignment, including all-versus-all sequence comparison \cite{FSA,MUSCLE}
+of multiple alignment, including all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}
 
@@
 Previous work showed that, for the purposes of estimating
-molecular evolutionary parameters
-such as the rates of insertion-deletion events \cite{Westesson2012-zg}
-or the ratio between synonymous abd nonsynonymous substitution rates \cite{Redelings},
-it is best to use an explicit probabilistic model of the sequence evolution process
+molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
+the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
+or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
 and to treat sequence alignment rigorously as a missing data estimation problem
@@
 estimation of indel rates, and that these biases are typically substantially worse
-at higher indel rates.
+at higher indel rates \citep{Westesson2012-zg}.
 
@@
 
-We also report the results of benchmarking ProtPal on the BAliBase 3.0 benchmark
+We also report an assessment of the structural alignment accuracy of IndelHistorian on the benchmark datasets BAliBase 3.0 \citep{ThompsonEtAl2005}
+and PREFAB 4.0 \citep{Edgar2010}
 
@@
 
-\cite{Westesson2012-zg}
+\citep{Westesson2012-zg}
 
@@
 Comparison to MCMC tools
-BaliPhy, PhylogenyCafe, HandAlign
+BaliPhy \citep{RedelingsSuchard2005,RedelingsSuchard2007,Redelings2014},
+StatAlign \citep{NovakEtAl2008,HermanEtAl2014},
+HandAlign \citep{WestessonBarquistHolmes2012}
 
@@
 
-Decision theory may also be useful in conjunction with MCMC tools \cite{HermanHeinEtc}
+Decision theory may also be useful to produce consensus alignments summarizing an MCMC run \citep{HermanEtAl2015}
 

### 2016-07-22 doc/main.tex
writing

@@
 
-\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: phylogenetic ancestral reconstruction}
+\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: fast and accurate ancestral sequence reconstruction using weighted finite-state transducers}
 \author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
@@
 
-Multiple sequence alignments are used for different purposes in bioinformatics,
+Synthesis of reconstructed ancestral proteins offers a direct way to test hypotheses of evolutionary adaptation \citep{UgaldeEtAl2004,OrtlundEtAl2007,GaucherEtAl2008}
+or to generate diversity in combinatorial libraries for bioengineering by directed evolution \citep{AlcolombriEtAl2011,SantiagoOrtizEtAl2015}.
+The computational workflow for such reconstructions typically involves building a multiple sequence alignment and then using statistical phylogenetics to reconstruct the substitution history
+\citep{Liberles2007}.
+However, it remains an open question whether the multiple alignment tools that perform best on the most commonly-used benchmarks
+(which generally use protein structure as a gold standard)
+are also the best alignment tools for performing ancestral sequence reconstruction.
+
+Multiple sequence alignments are used for several purposes in bioinformatics,
 and these different applications have led to different criteria for evaluating
-the accuracy of multiple alignment tools
-
+the accuracy of the tools.
 Homology modeling in protein structure prediction
 leads to accuracy metrics that quantify the number of correctly aligned residues
-compared to a benchmark dataset of structurally-informed reference alignments
-
-Significant progress has been made optimizing these metrics
-
-Successful techniques include using Bayesian decision theory to maximize the expected
+compared to a benchmark dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
+Techniques that have proven successful in this context include using Bayesian decision theory to maximize the expected
 score under the metrics \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
@@
 protein sequences under selection for folding, such as the reduced likelihood of indels in
-hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}
-
-PRANK\citep{LoytynojaGoldman2005}, Muscle \citep{Edgar2004b}, ClustalW\citep{LarkinEtAl2007}, MAFFT\citep{KatohEtAl2005}
-
-These have been accompanied by performance optimizations at the rate-limiting steps
-of multiple alignment, including all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}
+hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
+These have been accompanied by performance optimizations of the various steps involved in multiple alignment,
+the rate-limiting step generally being all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}.
 
 The study of sequence evolution and molecular phylogenetics gives rise to
-an alternative set of criteria for measuring alignment accuracy
-
+a different set of criteria for measuring alignment accuracy.
+As an example, for the purposes of predicting a protein structure,
+functionally equivalent sequences in a rapidly-evolving region (such as a loop)
+may be considered alignable even if they have been entirely replaced by a series of indel mutations
+(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
+but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.
 Previous work showed that, for the purposes of estimating
@@
 and to treat sequence alignment rigorously as a missing data estimation problem
-under this model
+under this model.
 
-The ProtPal program is one such approach which uses the theory of
-weighted finite-state transducers
-These are automata which can be multiplied like substitution matrices
-and so can be thought of as a strict generalization of Felsenstein's likelihood-phylogenetic
-approach to multiple sequence alignment
-(in fact, the use of transducers makes Felsenstein's pruning algorithm formally
-equivalent to a special case of Sankoff's multiple alignment algorithm)
+The ProtPal program \citep{Westesson2012-zg} is one such approach which uses the theory of
+weighted finite-state transducers,
+automata which can be multiplied together like substitution matrices \cite{BouchardCote2013}.
+This approach can be represented as a strict generalization of Felsenstein's celebrated
+algorithm for computing likelihoods on phylogenetic trees \citep{Felsenstein81}.
+In fact, the use of transducers makes Felsenstein's pruning algorithm formally
+equivalent to a special case of Sankoff's phylogenetic multiple alignment and RNA folding algorithm \citep{Sankoff85};
+the RNA folding component is included if one generalizes from finite-state transducers to pushdown automata \citep{BradleyHolmes2009}.
 
-A previously published benchmark of alignment methods on simulated sequence data
+Our earlier benchmark of alignment methods on simulated sequence data
 showed that most alignment algorithms introduce systematic biases into the
@@
 at higher indel rates \citep{Westesson2012-zg}.
-
 ProtPal was found to be the most accurate algorithm for indel rate estimation,
 followed by PRANK and then MUSCLE.
-
 However, the implementation of ProtPal published with that benchmark was too slow
-for practical use, and no benchmark of ProtPal on structural alignment benchmarks was available
+for practical use.
+Furthermore, no benchmark of ProtPal on structural alignment benchmarks was available,
+and---although we have noted that structure prediction is a different task---it would be
+useful to know just how different it is, according to the metrics.
 
 Here, we present a clean implementation of the algorithm underlying ProtPal
-in a new tool, Indel Historian
+in a new tool, Indel Historian.
 The tool is targeted to evolutionary applications such
-as ancestral sequence reconstruction and molecular phylogenetics
-
+as ancestral sequence reconstruction and molecular phylogenetics.
 We also report an assessment of the structural alignment accuracy of IndelHistorian on the benchmark datasets BAliBase 3.0 \citep{ThompsonEtAl2005}
-and PREFAB 4.0 \citep{Edgar2010}
+and PREFAB 4.0 \citep{Edgar2010}.
 

### 2016-07-22 doc/main.tex
writing

@@
 a different set of criteria for measuring alignment accuracy.
-As an example, for the purposes of predicting a protein structure,
-functionally equivalent sequences in a rapidly-evolving region (such as a loop)
-may be considered alignable even if they have been entirely replaced by a series of indel mutations
-(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
-but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.
-Previous work showed that, for the purposes of estimating
+Several independent studies have showed that, for the purposes of estimating
 molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
@@
 under this model.
+One conceptual explanation for this phenomenon is that, for the purposes of predicting a protein structure,
+functionally equivalent sequences in a rapidly-evolving region (such as a loop)
+may be considered alignable even if they have been entirely replaced by a series of indel mutations
+(e.g. the sequences $X$ and $Y$ in the mutation trajectory $AXB \to AXYB \to AYB$),
+but for the purposes of evolutionary analysis, one generally wants to count every indel event that has occurred.
 
@@
 Furthermore, no benchmark of ProtPal on structural alignment benchmarks was available,
-and---although we have noted that structure prediction is a different task---it would be
+and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
 useful to know just how different it is, according to the metrics.

### 2016-07-22 doc/main.tex
doc

--- 2016-07-22 doc/main.tex
@@
 
-\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: phylogenetic ancestral reconstruction}
+\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: fast and accurate ancestral sequence reconstruction}
 \author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
@@
 {\bf Motivation.}
+Ancestral reconstruction of indel histories is a specialized application of multiple sequence alignment.
+The alignment tool ProtPal, which uses weighted finite-state transducers to model indel evolution,
+out-performed other ancestral-reconstruction alignment tools in a previous simulation benchmark,
+but has been too slow for practical use.
 {\bf Results.}
+Indel Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
+On structural alignment benchmarks, it out-performs PRANK and is comparable to CLUSTALW.
 {\bf Availability and Implementation.}
 Indel Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
-It is written in C++11.
 {\bf Contact.}
@@
 
-Multiple sequence alignments are used for different purposes in bioinformatics,
-and these different applications have led to different criteria for evaluating
-the accuracy of multiple alignment tools
+Synthesis of reconstructed ancestral proteins offers a direct way to test hypotheses of evolutionary adaptation \citep{UgaldeEtAl2004,OrtlundEtAl2007,GaucherEtAl2008}
+or to generate diversity in combinatorial libraries for directed evolution \citep{AlcolombriEtAl2011,SantiagoOrtizEtAl2015}.
+A typical computational workflow involves building a multiple sequence alignment and then using statistical phylogenetics to reconstruct the substitution history
+\citep{Liberles2007}.
+However, it remains unclear whether the multiple alignment tools that perform best on common benchmarks
+(which generally use protein structure as a gold standard)
+are also the best tools for ancestral reconstruction.
 
-Homology modeling in protein structure prediction
-leads to accuracy metrics that quantify the number of correctly aligned residues
-compared to a benchmark dataset of structurally-informed reference alignments
+Multiple alignments are used for several purposes in bioinformatics,
+only one of which is homology-directed structure prediction,
+yet structurally-informed alignments have tended to dominate alignment benchmarks.
+Alignment accuracy scores (such as SPS, TCS, AMA, and the Cline shift score) quantify the number of correctly aligned residues
+compared to a benchmark dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
+Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
+alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
+and fine-tuning the alignment scoring function for known quirks of
+protein evolution, such as the reduced likelihood of indels in
+hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
+These have been accompanied by performance optimizations of 
+rate-limiting steps such as all-versus-all pairwise sequence comparison \citep{BradleyEtAl2009,Edgar2004b}.
 
-Significant progress has been made optimizing these metrics
+There is evidence that for evolutionary uses of alignments, such as reconstructing trees or ancestral sequences,
+different tools (or at least different tool-assessment metrics) might be required.
+Several independent studies have shown that, for the purposes of estimating
+molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
+dN/dS ratios \citep{Redelings2014},
+or trees \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
+and to treat sequence alignment rigorously as a statistical ``missing data'' problem.
+One possible explanation for this is that, for protein structure prediction,
+detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary,
+but evolutionary analyses can make more use of this information.
 
-Successful techniques include using Bayesian decision theory to maximize the expected
-score under the metrics \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
-and fine-tuning the alignment scoring function to reflect known details of
-protein sequences under selection for folding, such as the reduced likelihood of indels in
-hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}
-
-PRANK\citep{LoytynojaGoldman2005}, Muscle \citep{Edgar2004b}, ClustalW\citep{LarkinEtAl2007}, MAFFT\citep{KatohEtAl2005}
-
-These have been accompanied by performance optimizations at the rate-limiting steps
-of multiple alignment, including all-versus-all sequence comparison \citep{BradleyEtAl2009,Edgar2004b}
-
-The study of sequence evolution and molecular phylogenetics gives rise to
-an alternative set of criteria for measuring alignment accuracy
-
-Previous work showed that, for the purposes of estimating
-molecular evolutionary parameters---such as the rates of insertion-deletion events \citep{Westesson2012-zg},
-the ratio between synonymous and nonsynonymous substitution rates \citep{Redelings2014},
-or other evolutionary analyses \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
-and to treat sequence alignment rigorously as a missing data estimation problem
-under this model
-
-The ProtPal program is one such approach which uses the theory of
-weighted finite-state transducers
-These are automata which can be multiplied like substitution matrices
-and so can be thought of as a strict generalization of Felsenstein's likelihood-phylogenetic
-approach to multiple sequence alignment
-(in fact, the use of transducers makes Felsenstein's pruning algorithm formally
-equivalent to a special case of Sankoff's multiple alignment algorithm)
-
-A previously published benchmark of alignment methods on simulated sequence data
-showed that most alignment algorithms introduce systematic biases into the
-estimation of indel rates, and that these biases are typically substantially worse
-at higher indel rates \citep{Westesson2012-zg}.
-
+The ProtPal program \citep{Westesson2012-zg} is one such approach, modeling indel evolution using automata.
+weighted finite-state transducers,
+automata which can be multiplied together like substitution matrices \citep{BouchardCote2013},
+generalizing Felsenstein's celebrated
+algorithm for computing phylogenetic likelihoods from individual sites to whole sequences \citep{Felsenstein81}.
+%In fact, the use of transducers makes Felsenstein's pruning algorithm formally
+%equivalent to a special case of Sankoff's phylogenetic multiple alignment and RNA folding algorithm \citep{Sankoff85};
+%the RNA folding component is included if one generalizes from finite-state transducers to pushdown automata \citep{BradleyHolmes2009}.
+In an earlier benchmark of alignment methods on simulated sequence data,
 ProtPal was found to be the most accurate algorithm for indel rate estimation,
 followed by PRANK and then MUSCLE.
-
+Most alignment algorithms introduce systematic biases into the
+estimation of indel rates, and these biases are typically substantially worse
+at higher indel rates \citep{Westesson2012-zg}.
+These biases are minimized with ProtPal.
 However, the implementation of ProtPal published with that benchmark was too slow
-for practical use, and no benchmark of ProtPal on structural alignment benchmarks was available
+for practical use.
+Furthermore, no benchmark of ProtPal on structural alignment benchmarks has been available,
+and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
+useful to know just how different it is, according to the metrics.
 
 Here, we present a clean implementation of the algorithm underlying ProtPal
-in a new tool, Indel Historian
+in a new tool, Indel Historian.
 The tool is targeted to evolutionary applications such
-as ancestral sequence reconstruction and molecular phylogenetics
-
-We also report an assessment of the structural alignment accuracy of IndelHistorian on the benchmark datasets BAliBase 3.0 \citep{ThompsonEtAl2005}
-and PREFAB 4.0 \citep{Edgar2010}
+as ancestral sequence reconstruction and molecular phylogenetics.
+We also report an assessment of the alignment accuracy of IndelHistorian on structurally-informed benchmarks.
 
@@
 
-\citep{Westesson2012-zg}
+Indel Historian combines established algorithms from several sources.
+Like ProtPal, Indel Historian progressively climbs a tree from tips to root,
+using a finite-state transducer model of sequence evolution (similar to a Pair HMM)
+to build an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}.
+To find the guide tree, Indel Historian performs neighbor-joining \citep{SaitouNei87}
+on a guide alignment constructed via a sparse random subset of the all-vs-all pairwise alignment graph \citep{BradleyEtAl2009}.
+The progressive alignment step is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b} and optional MCMC \citep{WestessonBarquistHolmes2012}.
+Indel Historian also includes implementations of the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
+allowing substitution and indel rates to be estimated directly from sequence data.
 
@@
 
-Comparison to MCMC tools
-BaliPhy \citep{RedelingsSuchard2005,RedelingsSuchard2007,Redelings2014},
-StatAlign \citep{NovakEtAl2008,HermanEtAl2014},
-HandAlign \citep{WestessonBarquistHolmes2012}
-
 \section{Results}
 
-\section{Discussion}
+\begin{table}
+  \begin{tabular}{r|rr|rr}
+    & \multicolumn{2}{c|}{BAliBase 3.0} & \multicolumn{2}{c}{PREFAB 4} \\
+    & Mean SPS & Mean TCS & Mean SPS & Mean TCS \\
+    \hline
+IndelHistorian & 0.82 & 0.50 & 0.60 & 0.60 \\
+CLUSTALW & 0.79 & 0.45 & 0.62 & 0.62 \\
+PRANK & 0.73 & 0.35 & 0.51 & 0.51 \\
+MUSCLE & 0.89 & 0.64 & 0.73 & 0.73 \\
+  \end{tabular}
+  \caption{
+    Comparison of Indel Historian to other leading alignment programs using standard benchmarks
+    and the SPS/TCS scores \citep{ThompsonEtAl2005}.
+    Data for PRANK, CLUSTALW and MUSCLE are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+  }
+\end{table}
 
-The best-performing method MUSCLE introduces heuristics into its scoring scheme
-that improve accuracy.
+Supplementing the previously published data on the superior fidelity of the ProtPal algorithm
+at reconstructing indel histories \citep{Westesson2012-zg},
+Table~1 summarizes an evaluation of the ProtPal algorithm, as implemented in Indel Historian, on structural benchmarks.
+The comparison uses the BAliBase and PREFAB
+benchmarks, compiled partially or wholly using 3D structure information.
+In general, Indel Historian performs better than PRANK,
+a program which also aims for a phylogenetically meaningful alignment
+including ancestral sequences \citep{LoytynojaGoldman2008},
+and the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
+Compared to leading protein aligners that do not attempt ancestral reconstruction, Indel Historian performs better than CLUSTALW but worse than MUSCLE.
 
-It is harder to incorporate these sorts of modifications into a rigorous phylogenetic model
-primarily because they violate the assumption of independence
-between indel and substitution processes
-that most such models rely on
-
-Decision theory may also be useful to produce consensus alignments summarizing an MCMC run \citep{HermanEtAl2015}
+% The best-performing method MUSCLE introduces heuristics into its scoring scheme
+% that improve accuracy.
+% 
+% It is harder to incorporate these sorts of modifications into a rigorous phylogenetic model
+% primarily because they violate the assumption of independence
+% between indel and substitution processes
+% that most such models rely on
+% 
+% Comparison to MCMC tools
+% BaliPhy \citep{RedelingsSuchard2005,RedelingsSuchard2007,Redelings2014},
+% StatAlign \citep{NovakEtAl2008,HermanEtAl2014},
+% HandAlign \citep{WestessonBarquistHolmes2012}
+% 
+% Decision theory may also be useful to produce consensus alignments summarizing an MCMC run \citep{HermanEtAl2015}
 
@@
 
-Indel Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort), and the StackOverflow community.
+Indel Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), % Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort),
+and the StackOverflow community.
 

### 2016-07-22 doc/main.tex
final

@@
 {\bf Motivation.}
-Ancestral reconstruction of indel histories is a specialized application of multiple sequence alignment.
-The alignment tool ProtPal, which uses weighted finite-state transducers to model indel evolution,
-out-performed other ancestral-reconstruction alignment tools in a previous simulation benchmark,
+Reconstruction of indel histories and ancestral sequences is a specialized variation of multiple sequence alignment.
+The alignment tool ProtPal, based on a statistical phylogenetic model combining indel and substitution processes,
+reconstructed indel histories more accurately than other tools in an earlier simulation benchmark,
 but has been too slow for practical use.
@@
 Indel Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
-On structural alignment benchmarks, it out-performs PRANK and is comparable to CLUSTALW.
+Supplementing the earlier simulation results on fidelity of ancestral sequence reconstruction,
+evaluations on the structurally-informed datasets BAliBase and PREFAB are reported.
 {\bf Availability and Implementation.}
@@
 \citep{Liberles2007}.
-However, it remains unclear whether the multiple alignment tools that perform best on common benchmarks
+However, it is doubtful whether the multiple alignment tools that perform best on common benchmarks
 (which generally use protein structure as a gold standard)
@@
 only one of which is homology-directed structure prediction,
-yet structurally-informed alignments have tended to dominate alignment benchmarks.
+yet structurally-derived alignments have tended to dominate alignment benchmarks.
 Alignment accuracy scores (such as SPS, TCS, AMA, and the Cline shift score) quantify the number of correctly aligned residues
@@
 
-There is evidence that for evolutionary uses of alignments, such as reconstructing trees or ancestral sequences,
-different tools (or at least different tool-assessment metrics) might be required.
-Several independent studies have shown that, for the purposes of estimating
+There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
+different alignment tools (or at least different tool-assessment metrics) might be required.
+Studies have shown that, for the purposes of estimating
 molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
 dN/dS ratios \citep{Redelings2014},
-or trees \citep{LoytynojaGoldman2008}---it is best to use an explicit probabilistic model of the sequence evolution process
-and to treat sequence alignment rigorously as a statistical ``missing data'' problem.
-One possible explanation for this is that, for protein structure prediction,
+or trees \citep{LoytynojaGoldman2008}---it is best to use an explicit statistical model of the sequence evolution process
+and to treat alignment rigorously as a ``missing data'' problem.
+One possible explanation is that, for protein structure prediction,
 detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary,
-but evolutionary analyses can make more use of this information.
+while evolutionary analyses can make more use of this information.
 
-The ProtPal program \citep{Westesson2012-zg} is one such approach, modeling indel evolution using automata.
+The ProtPal program \citep{Westesson2012-zg} is one such approach, modeling indel evolution using 
 weighted finite-state transducers,
-automata which can be multiplied together like substitution matrices \citep{BouchardCote2013},
-generalizing Felsenstein's celebrated
-algorithm for computing phylogenetic likelihoods from individual sites to whole sequences \citep{Felsenstein81}.
+automata which can be multiplied together like substitution matrices \citep{BouchardCote2013}.
+%Transducers generalize Felsenstein's celebrated
+%algorithm for computing phylogenetic likelihoods from individual sites to whole sequences \citep{Felsenstein81}.
 %In fact, the use of transducers makes Felsenstein's pruning algorithm formally
@@
 In an earlier benchmark of alignment methods on simulated sequence data,
-ProtPal was found to be the most accurate algorithm for indel rate estimation,
+ProtPal was found to be the most accurate for indel rate estimation,
 followed by PRANK and then MUSCLE.
-Most alignment algorithms introduce systematic biases into the
+Most alignment tools introduce systematic biases into the
 estimation of indel rates, and these biases are typically substantially worse
-at higher indel rates \citep{Westesson2012-zg}.
+at higher rates \citep{Westesson2012-zg}.
 These biases are minimized with ProtPal.
@@
 for practical use.
-Furthermore, no benchmark of ProtPal on structural alignment benchmarks has been available,
-and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
-useful to know just how different it is, according to the metrics.
+%Furthermore, no benchmark of ProtPal on structural alignment benchmarks has been available,
+%and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
+%useful to know just how different it is, according to the metrics.
 
-Here, we present a clean implementation of the algorithm underlying ProtPal
+Here, we present a clean reimplementation of the algorithm underlying ProtPal
 in a new tool, Indel Historian.
-The tool is targeted to evolutionary applications such
-as ancestral sequence reconstruction and molecular phylogenetics.
-We also report an assessment of the alignment accuracy of IndelHistorian on structurally-informed benchmarks.
+%The tool is targeted to evolutionary applications such
+%as ancestral sequence reconstruction and molecular phylogenetics.
+We also report an assessment of the alignment accuracy %of IndelHistorian
+on structurally-informed benchmarks.
 
@@
 Like ProtPal, Indel Historian progressively climbs a tree from tips to root,
-using a finite-state transducer model of sequence evolution (similar to a Pair HMM)
-to build an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}.
-To find the guide tree, Indel Historian performs neighbor-joining \citep{SaitouNei87}
+building an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}.
+To find the guide tree, Indel Historian does neighbor-joining \citep{SaitouNei87}
 on a guide alignment constructed via a sparse random subset of the all-vs-all pairwise alignment graph \citep{BradleyEtAl2009}.
-The progressive alignment step is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b} and optional MCMC \citep{WestessonBarquistHolmes2012}.
-Indel Historian also includes implementations of the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
-allowing substitution and indel rates to be estimated directly from sequence data.
+Progressive alignment is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b} and optional MCMC \citep{WestessonBarquistHolmes2012}.
+Indel Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
+so substitution and indel rates can be estimated directly from sequence data.
 
@@
 at reconstructing indel histories \citep{Westesson2012-zg},
-Table~1 summarizes an evaluation of the ProtPal algorithm, as implemented in Indel Historian, on structural benchmarks.
+Table~1 summarizes an evaluation of Indel Historian
+on structural benchmarks.
 The comparison uses the BAliBase and PREFAB
@@
 In general, Indel Historian performs better than PRANK,
-a program which also aims for a phylogenetically meaningful alignment
-including ancestral sequences \citep{LoytynojaGoldman2008},
-and the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
+which also generates a phylogenetic ancestral reconstruction \citep{LoytynojaGoldman2008},
+and is the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
 Compared to leading protein aligners that do not attempt ancestral reconstruction, Indel Historian performs better than CLUSTALW but worse than MUSCLE.
+
+The benchmark did not include MCMC approaches
+such as BaliPhy \citep{Redelings2014}, StatAlign \citep{NovakEtAl2008} or HandAlign \citep{WestessonBarquistHolmes2012}.
+These are expected to be more accurate still, but take much longer to run.
 

### 2016-07-23 doc/main.tex
-refine, -nj default

@@
 \documentclass{bioinfo}
-\copyrightyear{2015} \pubyear{2015}
+\copyrightyear{2016} \pubyear{2016}
 
@@
 Reconstruction of indel histories and ancestral sequences is a specialized variation of multiple sequence alignment.
-The alignment tool ProtPal, based on a statistical phylogenetic model combining indel and substitution processes,
-reconstructed indel histories more accurately than other tools in an earlier simulation benchmark,
-but has been too slow for practical use.
+%The best alignment tools for ancestral reconstruction
+%are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
+In prior-published simulations,
+the alignment tool ProtPal (based on a statistical phylogenetic model combining indel and substitution processes)
+reconstructed indel histories more accurately than other tools,
+but has proven too slow for practical use.
 {\bf Results.}
@@
 
-Synthesis of reconstructed ancestral proteins offers a direct way to test hypotheses of evolutionary adaptation \citep{UgaldeEtAl2004,OrtlundEtAl2007,GaucherEtAl2008}
-or to generate diversity in combinatorial libraries for directed evolution \citep{AlcolombriEtAl2011,SantiagoOrtizEtAl2015}.
-A typical computational workflow involves building a multiple sequence alignment and then using statistical phylogenetics to reconstruct the substitution history
-\citep{Liberles2007}.
-However, it is doubtful whether the multiple alignment tools that perform best on common benchmarks
-(which generally use protein structure as a gold standard)
-are also the best tools for ancestral reconstruction.
-
 Multiple alignments are used for several purposes in bioinformatics,
-only one of which is homology-directed structure prediction,
-yet structurally-derived alignments have tended to dominate alignment benchmarks.
+only one of which is homology-directed protein structure prediction,
+yet structurally-derived alignments have tended to dominate benchmarks.
 Alignment accuracy scores (such as SPS, TCS, AMA, and the Cline shift score) quantify the number of correctly aligned residues
@@
 alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
-and fine-tuning the alignment scoring function for known quirks of
-protein evolution, such as the reduced likelihood of indels in
-hydrophobic regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
+and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
+hydrophobic protein regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
 These have been accompanied by performance optimizations of 
@@
 There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
-different alignment tools (or at least different tool-assessment metrics) might be required.
-Studies have shown that, for the purposes of estimating
+different alignment tools (or at least different tool-assessment metrics) might be preferable.
+Empirical studies suggest that, for the purposes of estimating
 molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
 dN/dS ratios \citep{Redelings2014},
-or trees \citep{LoytynojaGoldman2008}---it is best to use an explicit statistical model of the sequence evolution process
+or trees \citep{LoytynojaGoldman2008}---it is advantageous to use an explicit statistical model of the sequence evolution process
 and to treat alignment rigorously as a ``missing data'' problem.
 One possible explanation is that, for protein structure prediction,
-detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary,
-while evolutionary analyses can make more use of this information.
+detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
+(as structurally homologous loops can simply be aligned),
+whereas evolutionary analyses can make use of this information.
 
-The ProtPal program \citep{Westesson2012-zg} is one such approach, modeling indel evolution using 
-weighted finite-state transducers,
-automata which can be multiplied together like substitution matrices \citep{BouchardCote2013}.
-%Transducers generalize Felsenstein's celebrated
-%algorithm for computing phylogenetic likelihoods from individual sites to whole sequences \citep{Felsenstein81}.
-%In fact, the use of transducers makes Felsenstein's pruning algorithm formally
-%equivalent to a special case of Sankoff's phylogenetic multiple alignment and RNA folding algorithm \citep{Sankoff85};
-%the RNA folding component is included if one generalizes from finite-state transducers to pushdown automata \citep{BradleyHolmes2009}.
-In an earlier benchmark of alignment methods on simulated sequence data,
-ProtPal was found to be the most accurate for indel rate estimation,
-followed by PRANK and then MUSCLE.
-Most alignment tools introduce systematic biases into the
-estimation of indel rates, and these biases are typically substantially worse
-at higher rates \citep{Westesson2012-zg}.
-These biases are minimized with ProtPal.
-However, the implementation of ProtPal published with that benchmark was too slow
+One previous study sought to quantify systematic biases introduced into the
+estimation of indel rates \citep{Westesson2012-zg},
+using the evolution simulator indel-Seq-Gen \citep{StropeEtAl2009}.
+The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
+transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
+the least biases of the tools evaluated (RMS error 25\% for insertion rates and 25\% for deletion rates),
+followed by PRANK (RMS errors 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\% and 37\%).
+Biases were substantially worse at higher indel rates: e.g. for the highest simulated insertion rate of 0.08 insertions/substitution,
+the RMS estimation errors were 43\% (ProtPal), 64\% (PRANK) and 74\% (MUSCLE).
+
+Unfortunately, the implementation of ProtPal published with that benchmark was too slow
 for practical use.
-%Furthermore, no benchmark of ProtPal on structural alignment benchmarks has been available,
-%and---although we have noted that structure prediction is a different task from evolutionary analysis---it would be
-%useful to know just how different it is, according to the metrics.
-
 Here, we present a clean reimplementation of the algorithm underlying ProtPal
 in a new tool, Indel Historian.
-%The tool is targeted to evolutionary applications such
-%as ancestral sequence reconstruction and molecular phylogenetics.
-We also report an assessment of the alignment accuracy %of IndelHistorian
-on structurally-informed benchmarks.
+We also report an assessment of the alignment accuracy
+on structurally-derived benchmarks,
+complementing the above-described evolution-oriented simulation experiments.
 
@@
 Like ProtPal, Indel Historian progressively climbs a tree from tips to root,
-building an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}.
+building an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}
+using a time-dependent evolutionary model \citep{RivasEddy2015}.
 To find the guide tree, Indel Historian does neighbor-joining \citep{SaitouNei87}
-on a guide alignment constructed via a sparse random subset of the all-vs-all pairwise alignment graph \citep{BradleyEtAl2009}.
-Progressive alignment is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b} and optional MCMC \citep{WestessonBarquistHolmes2012}.
+on a guide alignment constructed from the all-vs-all pairwise alignment graph,
+or from a sparse random subset \citep{BradleyEtAl2009}.
+The guide alignment, which can also be supplied by another tool,
+can optionally constrain the progressive reconstruction,
+which is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b}.
 Indel Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
-so substitution and indel rates can be estimated directly from sequence data.
+so substitution and indel rates can be estimated directly from sequence data,
+summing over the most probable alignments.
 
@@
 
-\section{Results}
+\section{Results and Discussion}
 
@@
     and the SPS/TCS scores \citep{ThompsonEtAl2005}.
-    Data for PRANK, CLUSTALW and MUSCLE are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+    Data for CLUSTALW, PRANK (using +F option and CLUSTALW tree) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
   }
@@
 
-Supplementing the previously published data on the superior fidelity of the ProtPal algorithm
+Supplementing the previously published data on the fidelity of the ProtPal algorithm
 at reconstructing indel histories \citep{Westesson2012-zg},
 Table~1 summarizes an evaluation of Indel Historian
-on structural benchmarks.
-The comparison uses the BAliBase and PREFAB
-benchmarks, compiled partially or wholly using 3D structure information.
-In general, Indel Historian performs better than PRANK,
+on the BAliBase and PREFAB benchmarks,
+compiled partially or wholly using 3D structural alignments.
+In general, Indel Historian performs better on these structure-derived benchmarks than PRANK,
 which also generates a phylogenetic ancestral reconstruction \citep{LoytynojaGoldman2008},
 and is the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
-Compared to leading protein aligners that do not attempt ancestral reconstruction, Indel Historian performs better than CLUSTALW but worse than MUSCLE.
+Compared to leading protein aligners that do not attempt ancestral reconstruction,
+Indel Historian outperforms CLUSTALW \citep{LarkinEtAl2007} but not MUSCLE \citep{Edgar2004b},
+ProbCons \citep{DoEtAl2005} or MAFFT \citep{KatohEtAl2005}.
 
-The benchmark did not include MCMC approaches
+We did not benchmark MCMC approaches
 such as BaliPhy \citep{Redelings2014}, StatAlign \citep{NovakEtAl2008} or HandAlign \citep{WestessonBarquistHolmes2012}.
-These are expected to be more accurate still, but take much longer to run.
+These are expected to be more accurate still, but generally take much longer.
+MCMC samplers may be usefully supplemented by decision-theoretic approaches to summarize a sampling run \citep{HermanEtAl2015}.
+Other potential ways to improve accuracy include position-specific gap penalties, as used by MUSCLE \citep{Edgar2004b} and MAFFT \citep{KatohEtAl2005},
+and explicit modeling of tandem duplications \citep{SzalkowskiAnisimova2013}.
+All these approaches may be combined with additional information such as structural annotations to further improve alignments \citep{HermanEtAl2014}.
 

### 2016-07-23 src/recon.cpp

@@ -32,7 +32,7 @@ Reconstructor::Reconstructor()
     maxDistanceFromGuide (DefaultMaxDistanceFromGuide),
     tokenizeCodons (false),
     guideAlignTryAllPairs (true),
-    useUPGMA (true),
+    useUPGMA (false),
     jukesCantorDistanceMatrix (false),
     includeBestTraceInProfile (true),
     keepGapsOpen (false),
@@ -407,6 +407,11 @@ bool Reconstructor::parseSamplerArgs (deque<string>& argvec) {
       runMCMC = true;
       useUPGMA = true;
       argvec.pop_front();
+      argvec.pop_front();
+      return true;
+
+    } else if (arg == "-norefine") {
+      refineReconstruction = false;
       argvec.pop_front();
       return true;
 

### 2016-07-23 t/historian.cpp

@@ -128,7 +128,7 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "Following construction of the guide alignment, a tree is estimated using a\n"
     + "distance matrix method. By default this is UPGMA.\n"
     + "\n"
-    //    + "  -upgma          Use UPGMA to estimate tree (default)\n"
+    + "  -upgma          Use UPGMA to estimate tree (default for MCMC)\n"
     + "  -nj             Use neighbor-joining, not UPGMA, to estimate tree\n"
     + "  -jc             Use Jukes-Cantor-like estimates for distance matrix\n"
     + "\n"
@@ -233,6 +233,7 @@ int main (int argc, char** argv) {
 
     recon.reconstructRoot = true;
     recon.runMCMC = true;
+    recon.useUPGMA = true;
     recon.accumulateSubstCounts = false;
     recon.accumulateIndelCounts = false;
 

### 2016-07-23 doc/main.tex
revisions

@@
 the least biases of the tools evaluated (RMS error 25\% for insertion rates and 25\% for deletion rates),
-followed by PRANK (RMS errors 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\% and 37\%).
+followed by PRANK (RMS error 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\%, 37\%).
 Biases were substantially worse at higher indel rates: e.g. for the highest simulated insertion rate of 0.08 insertions/substitution,
@@
 Like ProtPal, Indel Historian progressively climbs a tree from tips to root,
-building an ancestral sequence profile that includes suboptimal alignments \citep{Westesson2012-zg}
+building an ancestral sequence profile that includes suboptimal alignments \citep{LeeGrassoSharlow2002,Westesson2012-zg}
 using a time-dependent evolutionary model \citep{RivasEddy2015}.
-To find the guide tree, Indel Historian does neighbor-joining \citep{SaitouNei87}
+To find the guide tree, Indel Historian does neighbor-joining % \citep{SaitouNei87}
 on a guide alignment constructed from the all-vs-all pairwise alignment graph,

### 2016-07-23 doc/main.tex
revisions

@@
 using a time-dependent evolutionary model \citep{RivasEddy2015}.
-To find the guide tree, Indel Historian does neighbor-joining % \citep{SaitouNei87}
+The guide tree is found by neighbor-joining % \citep{SaitouNei87}
 on a guide alignment constructed from the all-vs-all pairwise alignment graph,
-or from a sparse random subset \citep{BradleyEtAl2009}.
+or from a sparse random connected subgraph \citep{BradleyEtAl2009}.
 The guide alignment, which can also be supplied by another tool,
@@
 
-Supplementing the previously published data on the fidelity of the ProtPal algorithm
-at reconstructing indel histories \citep{Westesson2012-zg},
 Table~1 summarizes an evaluation of Indel Historian
@@
 compiled partially or wholly using 3D structural alignments.
+(This complements the previously published data on rate estimation:
+Indel Historian is primarily for evolutionary applications,
+but that should not preclude critical evaluation on a structural benchmark.)
 In general, Indel Historian performs better on these structure-derived benchmarks than PRANK,
-which also generates a phylogenetic ancestral reconstruction \citep{LoytynojaGoldman2008},
+which also performs ancestral reconstruction \citep{LoytynojaGoldman2008},
 and is the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
@@
 such as BaliPhy \citep{Redelings2014}, StatAlign \citep{NovakEtAl2008} or HandAlign \citep{WestessonBarquistHolmes2012}.
-These are expected to be more accurate still, but generally take much longer.
+These are expected to be more accurate, but generally take much longer.
 MCMC samplers may be usefully supplemented by decision-theoretic approaches to summarize a sampling run \citep{HermanEtAl2015}.

### 2016-07-23 doc/main.tex
revisions

@@
 and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
-hydrophobic protein regions \citep{KatohEtAl2005,Edgar2004b,LarkinEtAl2007}.
+hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
 These have been accompanied by performance optimizations of 
-rate-limiting steps such as all-versus-all pairwise sequence comparison \citep{BradleyEtAl2009,Edgar2004b}.
+rate-limiting steps such as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
 

### 2016-07-23 doc/main.tex
revisions

@@
     and the SPS/TCS scores \citep{ThompsonEtAl2005}.
-    Data for CLUSTALW, PRANK (using +F option and CLUSTALW tree) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
+    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
   }
@@
 and explicit modeling of tandem duplications \citep{SzalkowskiAnisimova2013}.
-All these approaches may be combined with additional information such as structural annotations to further improve alignments \citep{HermanEtAl2014}.
+Incorporating additional data such as structural annotations may further improve alignments \citep{HermanEtAl2014}.
 

### 2016-07-23 doc/main.tex
revisions

@@
 %are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
-In prior-published simulations,
+In previously published simulations,
 the alignment tool ProtPal (based on a statistical phylogenetic model combining indel and substitution processes)

### 2016-10-18 doc/sim/bin/add-indelign-params.js
bin/add-indelign-params.js

@@ -1,0 +1,6 @@
+#!/usr/bin/env node
+
+var fs = require("fs")
+var len = process.argv[2], maxIndelLen = process.argv[3], indelProb = process.argv[4], treeFile = process.argv[5]
+var tree = fs.readFileSync (treeFile, 'utf8')
+process.stdout.write ("[" + len + "]{" + maxIndelLen + "," + indelProb + "}" + tree)

### 2016-10-18 doc/sim/bin/dna-sim.js
bin/dna-sim.js

@@ -1,0 +1,20 @@
+#!/usr/bin/env node
+
+var fs = require("fs")
+var execSync = require("child_process").execSync
+function exec (cmd) {
+    console.log(cmd)
+    execSync(cmd)
+}
+
+var pow = process.argv[2], dir = process.argv[3], makefileFromDir = process.argv[4], indelProb = process.argv[5], reps = process.argv[6], treeFile = process.argv[7]
+
+var len = Math.pow (10, pow)
+var maxIndelLen = Math.pow (2, parseInt(pow) + 2)
+
+var paramFile = dir + "/params.isg"
+
+exec ("bin/add-indelign-params.js " + len + " " + maxIndelLen + " " + indelProb + " " + treeFile + " >" + paramFile)
+for (var rep = 1; rep <= reps; ++rep) {
+    exec ("cd " + dir + "; make -f " + makefileFromDir + " dna" + rep + ".hist.json")
+}

### 2016-10-18 doc/main.tex
Variable branch lengths

@@
 yet structurally-derived alignments have tended to dominate benchmarks.
-Alignment accuracy scores (such as SPS, TCS, AMA, and the Cline shift score) quantify the number of correctly aligned residues
-compared to a benchmark dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
-Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
-alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
-and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
-hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
-These have been accompanied by performance optimizations of 
-rate-limiting steps such as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
+There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
+different alignment tools (along with different tool-assessment metrics) might be preferable.
+Aligners that are optimized for detecting structural homology may not do so well at recovering information about substitution rates,
+whereas explicit statistical models of the sequence evolution process may do a better job at estimating these parameters.
 
-There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
-different alignment tools (or at least different tool-assessment metrics) might be preferable.
 Empirical studies suggest that, for the purposes of estimating
@@
 dN/dS ratios \citep{Redelings2014},
-or trees \citep{LoytynojaGoldman2008}---it is advantageous to use an explicit statistical model of the sequence evolution process
+or trees \citep{LoytynojaGoldman2008}---it is advantageous to use a statistical model of evolution
 and to treat alignment rigorously as a ``missing data'' problem.
@@
 
-One previous study sought to quantify systematic biases introduced into the
+The benchmarks commonly used to evaluate aligners use accuracy scores
+(such as SPS, TCS, AMA, and the Cline shift score)
+which quantify the number of correctly aligned residues
+compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
+Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
+alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
+and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
+hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
+Aligners have also been tuned for performance by optimizing rate-limiting steps
+uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
+
+In a previous study, we sought to quantify systematic biases introduced into the
 estimation of indel rates \citep{Westesson2012-zg},

### 2016-10-18 doc/sim/bin/add-indelign-params.js

@@ -1,6 +1,11 @@
 #!/usr/bin/env node
 
 var fs = require("fs")
-var len = process.argv[2], maxIndelLen = process.argv[3], indelProb = process.argv[4], treeFile = process.argv[5]
+var len = process.argv[2], maxIndelLen = process.argv[3], indelProb = process.argv[4], treeFile = process.argv[5], branchMultiplier = process.argv[6] || 1
 var tree = fs.readFileSync (treeFile, 'utf8')
+
+tree = tree.replace (/:([\d\.]+)/g, function (str, p1, offset, s) {
+    return ':' + (p1 * branchMultiplier)
+})
+
 process.stdout.write ("[" + len + "]{" + maxIndelLen + "," + indelProb + "}" + tree)

### 2016-10-18 doc/sim/bin/dna-sim.js
gp120da

@@ -16,5 +16,5 @@ var paramFile = dir + "/params.isg"
 
 exec ("bin/add-indelign-params.js " + len + " " + maxIndelLen + " " + indelProb + " " + treeFile + " >" + paramFile)
 for (var rep = 1; rep <= reps; ++rep) {
-    exec ("cd " + dir + "; make -f " + makefileFromDir + " dna" + rep + ".hist.json")
+    exec ("cd " + dir + "; make -f " + makefileFromDir + " dna" + rep + ".hist.stk")
 }

### 2016-10-18 doc/main.tex
writeup

@@
 \begin{table}
+  \begin{tabular}{r|r}
+    & RMS error \\
+    \hline
+Indel Historian & \\
+PRANK & \\
+ProbCons & \\
+MUSCLE & \\
+  \end{tabular}
+  \caption{
+    Comparison of Indel Historian to other alignment programs on a simulation benchmark
+    using datasets generated by indel-Seq-Gen \citep{StropeEtAl2009}
+    with parameters mimicking an alignment of HIV GP120 domains.
+  }
+\end{table}
+
+We first performed a simulation benchmark to assess Indel Historian's ability to reconstruct evolutionary parameters,
+compared to other tools.
+We based the simulation parameters on the evolutionary profile of the HIV GP120 envelope domain, as follows.
+We started with an alignment of ten GP120 domains from HIV and SIV and estimated a tree from this alignment.
+The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
+The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
+
+We further estimated the approximate indel rates from this alignment, and converted these to indel-Seq-Gen's parameter format.
+To one significant figure, the probability of opening an indel at any given site on a unit-length branch is $p=0.3$, and the sequence length is 500 amino acids.
+
+We then simulated 100 different alignments under these conditions, using indel-Seq-Gen, and attempted to estimate indel rates
+(a) from the true alignment; (b) using alignments or reconstructions created by ProbCons, MUSCLE, PRANK, and Indel Historian.
+In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
+compared to the true indel rates ($\lambda = \mu = \log(1-p)$).
+The RMS (root mean square) of these errors are reported in Table~1.
+
+\begin{table}
   \begin{tabular}{r|rr|rr}
@@
     \hline
-IndelHistorian & 0.82 & 0.50 & 0.60 & 0.60 \\
+Indel Historian & 0.82 & 0.50 & 0.60 & 0.60 \\
 CLUSTALW & 0.79 & 0.45 & 0.62 & 0.62 \\
 PRANK & 0.73 & 0.35 & 0.51 & 0.51 \\
-MUSCLE & 0.89 & 0.64 & 0.73 & 0.73 \\
-  \end{tabular}
+MAFFT & 0.87 & 0.61 & 0.69 & 0.69 \\
+ProbCons & 0.88 & 0.62 & 0.72 & 0.72 \\
+MUSCLE & 0.89 & 0.64 & 0.73 & 0.\end{tabular}
   \caption{
-    Comparison of Indel Historian to other leading alignment programs using standard benchmarks
+    Comparison of Indel Historian to other alignment programs using standard benchmarks
     and the SPS/TCS scores \citep{ThompsonEtAl2005}.
     Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
-    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree), MAFFT (v6.603), ProbCons (v1.12) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
   }
@@
 
-Table~1 summarizes an evaluation of Indel Historian
+Table~2 summarizes an evaluation of Indel Historian
 on the BAliBase and PREFAB benchmarks,
 compiled partially or wholly using 3D structural alignments.
-(This complements the previously published data on rate estimation:
-Indel Historian is primarily for evolutionary applications,
-but that should not preclude critical evaluation on a structural benchmark.)
+
 In general, Indel Historian performs better on these structure-derived benchmarks than PRANK,
@@
 ProbCons \citep{DoEtAl2005} or MAFFT \citep{KatohEtAl2005}.
+
+\begin{table}
+  \begin{tabular}{r|rr}
+    & \multicolumn{2}{c}{Average run time per alignment} \\
+ & Simulation & BAliBase 3 \\
+    \hline
+Indel Historian & & 55s \\
+PRANK & & 519s \\
+MUSCLE & & 1.9s \\
+\end{tabular}
+\caption{
+    Comparison of runtimes of Indel Historian, PRANK and MUSCLE on the simulation benchmark and the BAliBase 3 benchmark.
+  }
+\end{table}
+
+The results for the programs other than Indel Historian in Table~2 were taken from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+To make a direct comparison of runtimes, we re-ran the benchmarks for PRANK and MUSCLE on the same machine as the Indel Historian benchmark.
+These runtimes are summarized in Table~3: Indel Historian is an order of magnitude slower than MUSCLE, but an order of magnitude faster than PRANK.
 

### 2016-10-18 doc/main.tex
writeup

@@
     \hline
+True alignment & \\
 Indel Historian & \\
@@
 compared to the true indel rates ($\lambda = \mu = \log(1-p)$).
+
 The RMS (root mean square) of these errors are reported in Table~1.
+Note that even perfect knowledge of the true alignment does not guarantee perfect reconstruction of the underlying indel rate parameters,
+due to the possibility of indel events that are not directly observable (for example, when multiple indel events overlap on the same branch,
+they will generally be conflated into a single event by the inference algorithm).
 

### 2016-10-18 doc/main.tex
edits

@@
 ProbCons & 0.88 & 0.62 & 0.72 & 0.72 \\
-MUSCLE & 0.89 & 0.64 & 0.73 & 0.\end{tabular}
+MUSCLE & 0.89 & 0.64 & 0.73 & 0.73
+  \end{tabular}
   \caption{
@@
 Compared to leading protein aligners that do not attempt ancestral reconstruction,
-Indel Historian outperforms CLUSTALW \citep{LarkinEtAl2007} but not MUSCLE \citep{Edgar2004b},
-ProbCons \citep{DoEtAl2005} or MAFFT \citep{KatohEtAl2005}.
+Indel Historian outperforms CLUSTALW \citep{LarkinEtAl2007} on BaliBase3, but not MUSCLE \citep{Edgar2004b},
+ProbCons \citep{DoEtAl2005}, MAFFT \citep{KatohEtAl2005}, or CLUSTALW on PREFAB (though it is close).
 

### 2016-10-18 doc/main.tex
gp120tree.pdf

@@
 Indel Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
-Supplementing the earlier simulation results on fidelity of ancestral sequence reconstruction,
-evaluations on the structurally-informed datasets BAliBase and PREFAB are reported.
+Simulation results on fidelity of ancestral sequence reconstruction,
+along with evaluations on the structurally-informed datasets BAliBase and PREFAB, are reported.
 {\bf Availability and Implementation.}
@@
 
+\begin{figure}
+  \includegraphics[width=\columnwidth]{gp120tree.pdf}
+  \caption{
+    Tree of HIV and SIV GP120 domains that was used for the simulation benchmark.
+  }
+\end{figure}
+
 \begin{table}
@@
 We based the simulation parameters on the evolutionary profile of the HIV GP120 envelope domain, as follows.
-We started with an alignment of ten GP120 domains from HIV and SIV and estimated a tree from this alignment.
-The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
-The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
-
-We further estimated the approximate indel rates from this alignment, and converted these to indel-Seq-Gen's parameter format.
+We started with an alignment of ten GP120 domains from HIV and SIV and estimated a tree from this alignment (Figure~1).
+%The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
+%The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
+We used the tree to estimate the indel rates for the alignment, and converted these to indel-Seq-Gen's parameter format.
 To one significant figure, the probability of opening an indel at any given site on a unit-length branch is $p=0.3$, and the sequence length is 500 amino acids.

### 2016-10-19 doc/main.tex
edits

@@
   \includegraphics[width=\columnwidth]{gp120tree.pdf}
+%  \includegraphics[width=\columnwidth]{sym8tree.pdf}
   \caption{
-    Tree of HIV and SIV GP120 domains that was used for the simulation benchmark.
+    Tree of HIV and SIV GP120 domains, used for the simulation benchmark.
+%    Bottom: symmetric 8-taxon tree, also used for the simulation benchmark.
   }
@@
 \begin{table}
-  \begin{tabular}{r|r}
-    & RMS error \\
+  \begin{tabular}{r|rr}
+    & \multicolumn{2}{c}{GP120 tree} \\
+    & $\epsilon_\lambda$ & $\epsilon_\mu$ \\
     \hline
@@
 compared to the true indel rates ($\lambda = \mu = \log(1-p)$).
+We also repeated the benchmark with a symmetric 8-taxon binary tree whose 14 branches all had length 0.1.
 
-The RMS (root mean square) of these errors are reported in Table~1.
+The RMS (root mean square) errors
+$\epsilon_\lambda = \sqrt{\langle \left( \frac{\hat{\lambda}-\lambda}{\lambda} \right)^2 \rangle}$
+and
+$\epsilon_\mu = \sqrt{\langle \left( \frac{\hat{\mu}-\mu}{\mu} \right)^2 \rangle}$
+are reported in Table~1.
 Note that even perfect knowledge of the true alignment does not guarantee perfect reconstruction of the underlying indel rate parameters,
@@
     and the SPS/TCS scores \citep{ThompsonEtAl2005}.
-    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
-    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree), MAFFT (v6.603), ProbCons (v1.12) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
+%    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree), MAFFT (v6.603), ProbCons (v1.12) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
   }

### 2016-10-19 doc/sim/bin/dna-sim.js
improved dna sim

@@ -1,20 +1,0 @@
-#!/usr/bin/env node
-
-var fs = require("fs")
-var execSync = require("child_process").execSync
-function exec (cmd) {
-    console.log(cmd)
-    execSync(cmd)
-}
-
-var pow = process.argv[2], dir = process.argv[3], makefileFromDir = process.argv[4], indelProb = process.argv[5], reps = process.argv[6], treeFile = process.argv[7]
-
-var len = Math.pow (10, pow)
-var maxIndelLen = Math.pow (2, parseInt(pow) + 2)
-
-var paramFile = dir + "/params.isg"
-
-exec ("bin/add-indelign-params.js " + len + " " + maxIndelLen + " " + indelProb + " " + treeFile + " >" + paramFile)
-for (var rep = 1; rep <= reps; ++rep) {
-    exec ("cd " + dir + "; make -f " + makefileFromDir + " dna" + rep + ".hist.stk")
-}

### 2016-10-19 t/historian.cpp
helptext

@@ -84,7 +84,7 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "\n"
     + "  -profminpost <P>, -profsamples <N>\n"
     + "                  Specify minimum posterior prob. (P) for retaining DP states\n"
-    + "                   in profile (default " + TOSTRING(DefaultProfilePostProb) + "), or sample N states randomly\n"
+    + "                   in profile (default " + TOSTRING(DefaultProfilePostProb) + "), or sample N traces randomly\n"
     + "  -profmaxstates <S>\n"
     + "                  Limit profile to at most S states\n"
     + "                   (default of " + to_string(Reconstructor::defaultMaxProfileStates()) + " uses at most " + to_string((int) (.5+100*DefaultMaxDPMemoryFraction)) + "% of memory for DP matrix)\n"

### 2016-10-19 src/recon.cpp
Enable refinement by default

@@ -40,7 +40,7 @@ Reconstructor::Reconstructor()
     keepGapsOpen (false),
     usePosteriorsForProfile (true),
     reconstructRoot (true),
-    refineReconstruction (false),
+    refineReconstruction (true),
     accumulateSubstCounts (false),
     accumulateIndelCounts (false),
     predictAncestralSequence (false),

### 2016-10-19 src/recon.h

@@ -21,7 +21,7 @@
 
 #define AncestralSequencePostProbTag "PP"
 
-#define ReconFastAliasArgs {"-rndspan","-kmatchn","3","-band","10","-profmaxstates","1","-jc",}
+#define ReconFastAliasArgs {"-rndspan","-kmatchn","3","-band","10","-profmaxstates","1","-jc","-norefine"}
 
 class Reconstructor {
 public:

### 2016-10-19 t/historian.cpp

@@ -89,13 +89,18 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "                  Limit profile to at most S states\n"
     + "                   (default of " + to_string(Reconstructor::defaultMaxProfileStates()) + " uses at most " + to_string((int) (.5+100*DefaultMaxDPMemoryFraction)) + "% of memory for DP matrix)\n"
     + "\n"
+    + "  -norefine       Disable iterative refinement\n"
+    + "\n"
     + "Following alignment, ancestral sequence reconstruction can be performed.\n"
     + "\n"
     + "  -ancseq         Predict ancestral sequences (default is to leave them as " + Alignment::wildcardChar + "'s)\n"
     + "  -ancprob        Report posterior probabilities for ancestral residues\n"
     + "\n"
-    + "MCMC sampling allows for additional accuracy in historical reconstruction.\n"
+    + "For additional accuracy in historical reconstruction, the alignment can be\n"
+    + "iteratively refined, or MCMC-sampled. By default, refinement is enabled and\n"
+    + "MCMC is disabled.\n"
     + "\n"
+    + "  -norefine       Disable iterative refinement after initial reconstruction\n"
     + "  -mcmc           Run MCMC sampler after reconstruction\n"
     + "  -samples <N>    Number of MCMC iterations per sequence (default " + to_string(DefaultMCMCSamplesPerSeq) + ")\n"
     + "  -trace <file>   Specify MCMC trace filename\n"

### 2016-10-19 t/historian.cpp
Updated helptext

@@ -98,7 +98,7 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "\n"
     + "For additional accuracy in historical reconstruction, the alignment can be\n"
     + "iteratively refined, or MCMC-sampled. By default, refinement is enabled and\n"
-    + "MCMC is disabled.\n"
+    + "MCMC is disabled. (MCMC currently requires an ultrametric tree.)\n"
     + "\n"
     + "  -norefine       Disable iterative refinement after initial reconstruction\n"
     + "  -mcmc           Run MCMC sampler after reconstruction\n"
@@ -131,7 +131,7 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "  -kmatchoff      No kmer threshold, do full DP\n"
     + "\n"
     + "Following construction of the guide alignment, a tree is estimated using a\n"
-    + "distance matrix method. By default this is UPGMA.\n"
+    + "distance matrix method. By default this is neighbor-joining.\n"
     + "\n"
     + "  -upgma          Use UPGMA to estimate tree (default for MCMC)\n"
     + "  -nj             Use neighbor-joining, not UPGMA, to estimate tree\n"

### 2016-10-20 t/historian.cpp
more makefile targets for sim

@@ -89,8 +89,6 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "                  Limit profile to at most S states\n"
     + "                   (default of " + to_string(Reconstructor::defaultMaxProfileStates()) + " uses at most " + to_string((int) (.5+100*DefaultMaxDPMemoryFraction)) + "% of memory for DP matrix)\n"
     + "\n"
-    + "  -norefine       Disable iterative refinement\n"
-    + "\n"
     + "Following alignment, ancestral sequence reconstruction can be performed.\n"
     + "\n"
     + "  -ancseq         Predict ancestral sequences (default is to leave them as " + Alignment::wildcardChar + "'s)\n"
@@ -101,6 +99,7 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "MCMC is disabled. (MCMC currently requires an ultrametric tree.)\n"
     + "\n"
     + "  -norefine       Disable iterative refinement after initial reconstruction\n"
+    + "\n"
     + "  -mcmc           Run MCMC sampler after reconstruction\n"
     + "  -samples <N>    Number of MCMC iterations per sequence (default " + to_string(DefaultMCMCSamplesPerSeq) + ")\n"
     + "  -trace <file>   Specify MCMC trace filename\n"

### 2016-10-21 doc/sim/bin/add-indelign-params.js
Added .ma.tree.json for true tree estimates

@@ -1,11 +1,7 @@
 #!/usr/bin/env node
 
 var fs = require("fs")
-var len = process.argv[2], maxIndelLen = process.argv[3], indelProb = process.argv[4], treeFile = process.argv[5], branchMultiplier = process.argv[6] || 1
+var len = process.argv[2], maxIndelLen = process.argv[3], indelProb = process.argv[4], treeFile = process.argv[5]
 var tree = fs.readFileSync (treeFile, 'utf8')
-
-tree = tree.replace (/:([\d\.]+)/g, function (str, p1, offset, s) {
-    return ':' + (p1 * branchMultiplier)
-})
 
 process.stdout.write ("[" + len + "]{" + maxIndelLen + "," + indelProb + "}" + tree)

### 2016-10-21 doc/sim/bin/scale-tree.js

@@ -1,0 +1,11 @@
+#!/usr/bin/env node
+
+var fs = require("fs")
+var treeFile = process.argv[2], branchMultiplier = process.argv[3] || 1
+var tree = fs.readFileSync (treeFile, 'utf8')
+
+tree = tree.replace (/:([\d\.]+)/g, function (str, p1, offset, s) {
+    return ':' + (p1 * branchMultiplier)
+})
+
+process.stdout.write (tree)

### 2016-10-28 doc/sim/bin/results.js
Summary

@@ -1,0 +1,63 @@
+#!/usr/bin/env node
+
+var fs = require("fs");
+
+var binsize = process.argv[2],
+    binmax = process.argv[3],
+    nbins = (binmax/binsize)+1,
+    togetherFlag = eval(process.argv[4]),
+    methods = process.argv[5].split(" "),
+    dirs = process.argv.slice(6)
+
+function incbin(r,bins) {
+    var bin = Math.round(r/binsize)
+    bins[bin] = (bins[bin] || 0) + 1
+}
+
+var binStart = Array.apply(null,Array(nbins)).map (function(x,n) { return (n*binsize).toPrecision(2) })
+function printbins(label,headers,bins) {
+    binStart.forEach (function (start, n) {
+        process.stdout.write ([label]
+                              .concat(headers)
+                              .concat([start,bins[n] || 0,"\n"])
+                              .join(" "))
+    })
+}
+
+process.stdout.write (["event","method"]
+                      .concat (togetherFlag ? [] : ["rate","tree","multiplier"])
+                      .concat(["bin","count","\n"])
+                      .join (" "))
+
+methods.forEach (function(method) {
+    var ibins = {}, dbins = {}
+    dirs.forEach (function(dir) {
+        var match = /.*rate([\d\.]+)\/([a-z]+)(\d+)/.exec(dir),
+            rate = match[1],
+            tree = match[2],
+            mul = match[3],
+            filenameRegex = new RegExp('\\d+\\.' + method + '\\.json'),
+            headers
+        fs.readdirSync(dir)
+            .filter (function (filename) { return filenameRegex.test(filename) })
+            .map (function (filename) {
+                try {
+                    var json = JSON.parse(fs.readFileSync(dir+"/"+filename));
+                    incbin(json.insrate/rate,ibins);
+                    incbin(json.delrate/rate,dbins)
+                } catch(err) {
+                    process.stderr.write(err.toString()+" ("+dir+"/"+filename+")\n")
+                }
+            })
+        if (!togetherFlag) {
+            printbins("ins",[method,rate,tree,mul],ibins)
+            printbins("del",[method,rate,tree,mul],dbins)
+            ibins = {}
+            dbins = {}
+        }
+    })
+    if (togetherFlag) {
+        printbins("ins",[method],ibins)
+        printbins("del",[method],dbins)
+    }
+})

### 2016-10-28 doc/sim/bin/plotresults.R
bin/plotresults.R

@@ -1,0 +1,20 @@
+install.packages("ggplot2")
+install.packages("gridExtra")
+library(ggplot2)
+library(gridExtra)
+
+data <- read.csv(file="gp120sim/together.1-100.dat",sep=" ")
+
+maIns <- ggplot(data[data$method=="ma.tree"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+prankIns <- ggplot(data[data$method=="prank"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+histIns <- ggplot(data[data$method=="hist"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+muscleIns <- ggplot(data[data$method=="muscle"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+probIns <- ggplot(data[data$method=="probcons"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+
+maDel <- ggplot(data[data$method=="ma.tree"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
+prankDel <- ggplot(data[data$method=="prank"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
+histDel <- ggplot(data[data$method=="hist"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
+muscleDel <- ggplot(data[data$method=="muscle"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
+probDel <- ggplot(data[data$method=="probcons"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
+
+grid.arrange(maIns,maDel,histIns,histDel,prankIns,prankDel,muscleIns,muscleDel,probIns,probDel,ncol=2)

### 2016-10-28 doc/sim/bin/results.js

@@ -16,17 +16,21 @@ function incbin(r,bins) {
 
 var binStart = Array.apply(null,Array(nbins)).map (function(x,n) { return (n*binsize).toPrecision(2) })
 function printbins(label,headers,bins) {
+    var total = 0
+    Object.keys(bins).forEach (function (n) {
+        total += bins[n]
+    })
     binStart.forEach (function (start, n) {
         process.stdout.write ([label]
                               .concat(headers)
-                              .concat([start,bins[n] || 0,"\n"])
-                              .join(" "))
+                              .concat([start,(bins[n] || 0) / total])
+                              .join(" ") + "\n")
     })
 }
 
 process.stdout.write (["event","method"]
                       .concat (togetherFlag ? [] : ["rate","tree","multiplier"])
-                      .concat(["bin","count","\n"])
+                      .concat(["bin","count\n"])
                       .join (" "))
 
 methods.forEach (function(method) {

### 2016-10-28 doc/sim/bin/plotresults.R
bin/results.js

@@ -17,4 +17,5 @@ histDel <- ggplot(data[data$method=="hist"&data$event=="del",],aes(x=bin,weight=
 muscleDel <- ggplot(data[data$method=="muscle"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
 probDel <- ggplot(data[data$method=="probcons"&data$event=="del",],aes(x=bin,weight=count))+geom_bar()
 
-grid.arrange(maIns,maDel,histIns,histDel,prankIns,prankDel,muscleIns,muscleDel,probIns,probDel,ncol=2)
+g <- arrangeGrob(maIns,maDel,histIns,histDel,prankIns,prankDel,muscleIns,muscleDel,probIns,probDel,ncol=2)
+ggsave("together.pdf",g)

### 2016-10-28 doc/sim/bin/plotresults.R
Makefile

@@ -3,7 +3,9 @@ install.packages("gridExtra")
 library(ggplot2)
 library(gridExtra)
 
-data <- read.csv(file="gp120sim/together.1-100.dat",sep=" ")
+data.raw <- read.csv(file="gp120sim/results.1-100.dat",sep=" ")
+
+data <- data.raw[data$tree="tree"]
 
 maIns <- ggplot(data[data$method=="ma.tree"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
 prankIns <- ggplot(data[data$method=="prank"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()

### 2016-10-28 doc/sim/bin/summary.js

@@ -1,0 +1,31 @@
+#!/usr/bin/env node
+var fs = require("fs");
+var methods = process.argv[2].split(" ");
+var dirs = process.argv.slice(3);
+function mean_rmserr(n,m1,m2){m = m1/n;return [m.toPrecision(6),Math.sqrt(m2/n-2*m+1).toPrecision(6)]}
+methods.forEach (function (method) {
+    var n = 0, m1i = 0, m2i = 0, m1d = 0, m2d = 0;
+    dirs.forEach (function (dir) {
+        var match = /.*rate([\d\.]+).*/.exec(dir);
+        var rate = match[1];
+        var filenameRegex = new RegExp('\\d+\\.' + method + '\\.json');
+        fs.readdirSync(dir)
+            .filter (function (filename) { return filenameRegex.test(filename) })
+            .map (function (filename) {
+                try {
+                    var json = JSON.parse(fs.readFileSync(dir+'/'+filename));
+                    var ir = json.insrate/rate;
+                    var dr = json.delrate/rate;
+                    m1i += ir;
+                    m2i += ir*ir;
+                    m1d += dr;
+                    m2d += dr*dr;
+                    ++n;
+                } catch (err) {
+                    process.stderr.write(err.toString()+" ("+dir+"/"+filename+")\n")
+                }
+            })
+    })
+    if (n > 0)
+        process.stdout.write([method].concat(mean_rmserr(n,m1i,m2i)).concat(mean_rmserr(n,m1d,m2d)).join(" ")+"\n")
+})

### 2016-10-30 doc/sim/bin/plotresults.R
bench/Makefile

@@ -1,8 +1,12 @@
+#!/usr/bin/env R
+
 install.packages("ggplot2")
 install.packages("gridExtra")
 library(ggplot2)
 library(gridExtra)
 
+rms.sym <- read.csv(file="gp120sim/allsym.1-100.summary",sep=" ")
+rms.tree <- read.csv(file="gp120sim/alltree.1-100.summary",sep=" ")
 data.raw <- read.csv(file="gp120sim/results.1-100.dat",sep=" ")
 
 data <- data.raw[data$tree="tree"]

### 2016-10-30 doc/sim/bin/summary.js

@@ -3,6 +3,7 @@ var fs = require("fs");
 var methods = process.argv[2].split(" ");
 var dirs = process.argv.slice(3);
 function mean_rmserr(n,m1,m2){m = m1/n;return [m.toPrecision(6),Math.sqrt(m2/n-2*m+1).toPrecision(6)]}
+process.stdout.write(["method","insmean","insrmse","delmean","delrmse","insdelmean","insdelrmse"].join(" ")+"\n")
 methods.forEach (function (method) {
     var n = 0, m1i = 0, m2i = 0, m1d = 0, m2d = 0;
     dirs.forEach (function (dir) {
@@ -27,5 +28,9 @@ methods.forEach (function (method) {
             })
     })
     if (n > 0)
-        process.stdout.write([method].concat(mean_rmserr(n,m1i,m2i)).concat(mean_rmserr(n,m1d,m2d)).join(" ")+"\n")
+        process.stdout.write([method]
+                             .concat(mean_rmserr(n,m1i,m2i))
+                             .concat(mean_rmserr(n,m1d,m2d))
+                             .concat(mean_rmserr(n*2,m1i+m1d,m2i+m2d))
+                             .join(" ")+"\n")
 })

### 2016-10-30 doc/sim/bin/plotresults.R
results.pdf

@@ -1,27 +1,39 @@
 #!/usr/bin/env R
 
-install.packages("ggplot2")
-install.packages("gridExtra")
+# install.packages("ggplot2")
+# install.packages("gridExtra")
 library(ggplot2)
 library(gridExtra)
 
+rms.all <- read.csv(file="gp120sim/all.1-100.summary",sep=" ")
 rms.sym <- read.csv(file="gp120sim/allsym.1-100.summary",sep=" ")
 rms.tree <- read.csv(file="gp120sim/alltree.1-100.summary",sep=" ")
-data.raw <- read.csv(file="gp120sim/results.1-100.dat",sep=" ")
 
-data <- data.raw[data$tree="tree"]
+data.all <- read.csv(file="gp120sim/results.1-100.dat",sep=" ")
+data.sym <- data.all[data.all$tree=="sym",]
+data.tree <- data.all[data.all$tree=="tree",]
 
-maIns <- ggplot(data[data$method=="ma.tree"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
-prankIns <- ggplot(data[data$method=="prank"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
-histIns <- ggplot(data[data$method=="hist"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
-muscleIns <- ggplot(data[data$method=="muscle"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
-probIns <- ggplot(data[data$method=="probcons"&data$event=="ins",],aes(x=bin,weight=count))+geom_bar()
+simplot <- function(method,title) {
+	rms <- rms.tree
+	data <- data.tree
+	err.mean <- rms$insdelmean[rms$method==method]
+	err.rms <- rms$insdelrmse[rms$method==method]
+	return (ggplot(data[data$method==method,],aes(x=bin,weight=count))
+		+geom_bar()
+		+annotate("text",label=paste0(title,"\nMean error: ",sprintf("%.2f",round(err.mean-1,digits=2)),"\nRMS error: ",sprintf("%.2f",round(err.rms,digits=2))),x=1.6,y=1.75)
+		+xlab("Estimated rate / true rate")
+		+scale_x_continuous(limits = c(0.25,1.75))
+		+scale_y_continuous(limits = c(0,3))
+		+theme(axis.title.y=element_blank(),
+		       axis.text.y=element_blank(),
[…]

### 2016-10-30 doc/main.tex
Revised paper

--- 2016-10-18 doc/main.tex
@@
 
-\title[Indel Historian: phylogenetic ancestral reconstruction]{Indel Historian: fast and accurate ancestral sequence reconstruction}
+\title[Historian: accurate reconstruction of sequences and rates]{Historian: accurate reconstruction of ancestral sequences and evolutionary rates}
 \author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
@@
 {\bf Motivation.}
-Reconstruction of indel histories and ancestral sequences is a specialized variation of multiple sequence alignment.
+Reconstruction of ancestral sequence histories,
+and estimation of parameters like indel rates,
+are improved by using explicit evolutionary models
+and summing over uncertain alignments.
 %The best alignment tools for ancestral reconstruction
 %are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
-In previously published simulations,
-the alignment tool ProtPal (based on a statistical phylogenetic model combining indel and substitution processes)
-reconstructed indel histories more accurately than other tools,
-but has proven too slow for practical use.
+The previous best tool for this purpose was ProtPal,
+but this tool was too slow for practical use.
 {\bf Results.}
-Indel Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
+Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
 Simulation results on fidelity of ancestral sequence reconstruction,
-along with evaluations on the structurally-informed datasets BAliBase and PREFAB, are reported.
+along with evaluations on the structurally-informed dataset BAliBase 3.0,
+recommend Historian for evolutionary applications.
 {\bf Availability and Implementation.}
-Indel Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
+Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
 {\bf Contact.}
@@
 only one of which is homology-directed protein structure prediction,
-yet structurally-derived alignments have tended to dominate benchmarks.
-There is evidence that for evolutionary applications, such as reconstructing trees or ancestral sequences,
-different alignment tools (along with different tool-assessment metrics) might be preferable.
+yet this is exactly the application that has tended to dominate alignment benchmarks.
+For evolutionary applications, such as reconstructing trees or ancestral sequences,
+there is evidence that
+different alignment tools (along with different tool-assessment metrics) might be preferable \citep{LoytynojaGoldman2008,Westesson2012-zg}.
 Aligners that are optimized for detecting structural homology may not do so well at recovering information about substitution rates,
 whereas explicit statistical models of the sequence evolution process may do a better job at estimating these parameters.
+Furthermore, the ideal way to estimate evolutionary parameters is not to use a single point estimate of the alignment, but to sum over alignments as a ``nuisance variable''.
+Almost no tools do this, with the exception of MCMC samplers \citep{WestessonBarquistHolmes2012,Redelings2014}.
 
@@
 and to treat alignment rigorously as a ``missing data'' problem.
-One possible explanation is that, for protein structure prediction,
-detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
-(as structurally homologous loops can simply be aligned),
-whereas evolutionary analyses can make use of this information.
-
-The benchmarks commonly used to evaluate aligners use accuracy scores
-(such as SPS, TCS, AMA, and the Cline shift score)
-which quantify the number of correctly aligned residues
-compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
-Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
-alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
-and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
-hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
-Aligners have also been tuned for performance by optimizing rate-limiting steps
-uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
-
+%One possible explanation is that, for protein structure prediction,
+%detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
+%(as structurally homologous loops can simply be aligned),
+%whereas evolutionary analyses can make use of this information.
+%
+% The benchmarks commonly used to evaluate aligners use accuracy scores
+% (such as SPS, TCS, AMA, and the Cline shift score)
+% which quantify the number of correctly aligned residues
+% compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
+% Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
+% alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
+% and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
+% hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
+% Aligners have also been tuned for performance by optimizing rate-limiting steps
+% uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
+% 
 In a previous study, we sought to quantify systematic biases introduced into the
-estimation of indel rates \citep{Westesson2012-zg},
-using the evolution simulator indel-Seq-Gen \citep{StropeEtAl2009}.
+estimation of indel rates, using a simulation benchmark \citep{Westesson2012-zg}.
 The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
 transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
-the least biases of the tools evaluated (RMS error 25\% for insertion rates and 25\% for deletion rates),
-followed by PRANK (RMS error 40\% for insertions, 34\% for deletions) and then a MUSCLE-based workflow (45\%, 37\%).
-Biases were substantially worse at higher indel rates: e.g. for the highest simulated insertion rate of 0.08 insertions/substitution,
-the RMS estimation errors were 43\% (ProtPal), 64\% (PRANK) and 74\% (MUSCLE).
-
+the least biases of the tools evaluated.
 Unfortunately, the implementation of ProtPal published with that benchmark was too slow
 for practical use.
+
 Here, we present a clean reimplementation of the algorithm underlying ProtPal
-in a new tool, Indel Historian.
-We also report an assessment of the alignment accuracy
+in a new tool, Historian, that can also estimate rates by summing over alignments.
+We report an assessment of the alignment accuracy
 on structurally-derived benchmarks,
-complementing the above-described evolution-oriented simulation experiments.
+together with a simulation-based experiment to quantify the accuracy of indel rate estimation.
 
@@
 
-Indel Historian combines established algorithms from several sources.
-Like ProtPal, Indel Historian progressively climbs a tree from tips to root,
+Historian combines established algorithms from several sources.
+Like ProtPal, Historian progressively climbs a tree from tips to root,
 building an ancestral sequence profile that includes suboptimal alignments \citep{LeeGrassoSharlow2002,Westesson2012-zg}
@@
 can optionally constrain the progressive reconstruction,
-which is followed by iterative refinement \citep{HolmesBruno2001,Edgar2004b}.
-Indel Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
-so substitution and indel rates can be estimated directly from sequence data,
-summing over the most probable alignments.
+which is followed by iterative refinement. % \citep{HolmesBruno2001,Edgar2004b}.
+
+Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
+so substitution and indel rates can be estimated directly from sequence data.
+Since the method builds an HMM that captures suboptimal as well as optimal alignments---rather
+like a partial order alignment \citep{LeeGrassoSharlow2002}---the program
+can also estimate rates in an ``alignment-free'' way (i.e. summing over alignments)
+simply by running the Forward-Backward algorithm on this HMM,
+and using the posterior counts to weight the phylo-EM updates.
 
@@
   \caption{
-    Tree of HIV and SIV GP120 domains that was used for the simulation benchmark.
+    Tree of selected HIV and SIV GP120 domains, used for the simulation benchmark.
   }
@@
 
+\begin{figure}
+  \includegraphics[width=\columnwidth]{sim/results.pdf}
+  \caption{
+    Results of the simulation benchmark.
+    For each method evaluated, the distribution of the ratio of inferred to true
+    indel rate is shown, along with the mean and root-mean-squared error for this
+    ratio (which, for a perfect rate inference, would be equal to 1).
+  }
+\end{figure}
+
+We first performed a simulation benchmark to assess Historian's ability to reconstruct evolutionary parameters,
+compared to other tools.
+We based the simulation parameters on the evolutionary profile of the HIV/SIV GP120 envelope domain, as follows.
+We started with a tree made from an alignment of ten HIV and SIV GP120 domains (Figure~1).
+%The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
+%The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
+We used the tree to estimate the indel rates for the alignment,
+and simulated 100 different alignments on this tree
+using a sequence length of 500 amino acids
+(roughly the same as the GP120 domain)
+with the expected substitution rate ($\sigma$) set to one substitution per unit of time
+and the simulated insertion and deletion rates ($\lambda,\mu$)
+set to the midpoint of the insertion and deletion rates
+estimated for GP120 (0.028 indels per unit of time).
+For a substitution model, we used the Dayhoff PAM matrix,
+which we judged not to give an unfair advantage either to Historian
+(which by default uses a matrix estimated from PFAM)
+or to Prank (which uses the WAG matrix).
+
+From the simulated sequence data, we estimated indel rates
+(a) using the true evolutionary alignment (supplied to Historian);
+(b) using Historian (in alignment-free mode);
+and
+(c,d,e) using Prank, ProbCons and Muscle,
+the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
+In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
+compared to the simulated indel rates.
+Having performed this experiment
+using the rates estimated for GP120 ($\sigma=1,\lambda=\mu=0.028$),
+we repeated the simulation, varying the rate parameters
+($\sigma=2,\lambda=\mu=0.056;
+\sigma=5,\lambda=\mu=0.14;
+\sigma=2,\lambda=\mu=0.112;
+\sigma=5,\lambda=\mu=0.28;
+\sigma=5,\lambda=\mu=0.005;
+\sigma=5,\lambda=\mu=0.01$)
+while still using the tree of Figure~1.
+
+The results of this experiment are summarized in Figure~2.
+Note that even perfect knowledge of the true evolutionary alignment does not guarantee perfect reconstruction of the underlying indel rate parameters.
+Indel events can overlap and thus be under-counted,
+leading to a small negative bias in the estimated rate,
+and stochastic noise from a finite sample leads to a spread in the distribution
+of estimated rates from individual alignments.
+Historian and Prank (which both explicitly aim to provide ancestral reconstructions)
+are the most accurate methods for rate estimation,
+with a slight edge over ProbCons and Muscle.
+
+As well as varying the rate parameters,
+we also repeated the benchmark on a symmetric 8-taxon binary tree.
+This simulation yielded similar patterns:
+Historian and Prank are still comparable (with a slight edge to Prank in the
+symmetric tree, vs Historian in the GP120 tree),
+followed by ProbCons, followed by Muscle.
+
 \begin{table}
-  \begin{tabular}{r|r}
-    & RMS error \\
+  \begin{tabular}{r|ccc}
+    & Mean SPS & Mean TCS & Notes \\
     \hline
-True alignment & \\
-Indel Historian & \\
-PRANK & \\
-ProbCons & \\
-MUSCLE & \\
+Historian v1.0 & 0.82 & 0.50 & \\
+ClustalW & 0.79 & 0.45 & From {\tt drive5.com} \\
+Prank v.100701 & 0.78 & 0.44 \\
+%MAFFT & 0.87 & 0.61 \\
+ProbCons v1.12 & 0.88 & 0.62 & From {\tt drive5.com} \\
+Muscle v3.8.31 & 0.89 & 0.64 & 
   \end{tabular}
   \caption{
-    Comparison of Indel Historian to other alignment programs on a simulation benchmark
-    using datasets generated by indel-Seq-Gen \citep{StropeEtAl2009}
-    with parameters mimicking an alignment of HIV GP120 domains.
+    Comparison of Historian to other alignment programs using the BAliBase 3.0 benchmark
+    with SPS and TCS alignment quality scores \citep{ThompsonEtAl2005}.
+%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
+%    Data for ClustalW, Prank (+F option, ClustalW tree), MAFFT (v6.603), ProbCons (v1.12) and Muscle (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
   }
@@
 
-We first performed a simulation benchmark to assess Indel Historian's ability to reconstruct evolutionary parameters,
-compared to other tools.
-We based the simulation parameters on the evolutionary profile of the HIV GP120 envelope domain, as follows.
-We started with an alignment of ten GP120 domains from HIV and SIV and estimated a tree from this alignment (Figure~1).
-%The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
-%The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
-We used the tree to estimate the indel rates for the alignment, and converted these to indel-Seq-Gen's parameter format.
-To one significant figure, the probability of opening an indel at any given site on a unit-length branch is $p=0.3$, and the sequence length is 500 amino acids.
-
-We then simulated 100 different alignments under these conditions, using indel-Seq-Gen, and attempted to estimate indel rates
-(a) from the true alignment; (b) using alignments or reconstructions created by ProbCons, MUSCLE, PRANK, and Indel Historian.
-In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
-compared to the true indel rates ($\lambda = \mu = \log(1-p)$).
-
-The RMS (root mean square) of these errors are reported in Table~1.
-Note that even perfect knowledge of the true alignment does not guarantee perfect reconstruction of the underlying indel rate parameters,
-due to the possibility of indel events that are not directly observable (for example, when multiple indel events overlap on the same branch,
-they will generally be conflated into a single event by the inference algorithm).
+Table~2 summarizes an evaluation of Historian
+on the BAliBase 3.0 benchmark,
+compiled using 3D structural alignments.
+%
+In general, Historian performs better on these structure-derived benchmarks than Prank,
+which also performs ancestral reconstruction \citep{LoytynojaGoldman2008}.
+Compared to leading protein aligners that do not attempt ancestral reconstruction,
+Historian outperforms ClustalW \citep{LarkinEtAl2007}, but not Muscle \citep{Edgar2004b}
+or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.
 
 \begin{table}
-  \begin{tabular}{r|rr|rr}
-    & \multicolumn{2}{c|}{BAliBase 3.0} & \multicolumn{2}{c}{PREFAB 4} \\
-    & Mean SPS & Mean TCS & Mean SPS & Mean TCS \\
+  \begin{tabular}{r|c}
+    & Average run time per BAliBase 3.0 alignment \\
     \hline
-Indel Historian & 0.82 & 0.50 & 0.60 & 0.60 \\
-CLUSTALW & 0.79 & 0.45 & 0.62 & 0.62 \\
-PRANK & 0.73 & 0.35 & 0.51 & 0.51 \\
-MAFFT & 0.87 & 0.61 & 0.69 & 0.69 \\
-ProbCons & 0.88 & 0.62 & 0.72 & 0.72 \\
-MUSCLE & 0.89 & 0.64 & 0.73 & 0.73
+Historian v1.0 & 55s \\
+Prank v.100701 & 520s \\
+Muscle v3.8.31 & 1.9s \\
 \end{tabular}
 \caption{
-    Comparison of Indel Historian to other alignment programs using standard benchmarks
-    and the SPS/TCS scores \citep{ThompsonEtAl2005}.
-    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
-    Data for CLUSTALW, PRANK (+F option, CLUSTALW tree), MAFFT (v6.603), ProbCons (v1.12) and MUSCLE (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
+    Comparison of runtimes of Historian, Prank and Muscle on the BAliBase 3.0 benchmark.
   }
@@
 
-Table~2 summarizes an evaluation of Indel Historian
-on the BAliBase and PREFAB benchmarks,
-compiled partially or wholly using 3D structural alignments.
-
-In general, Indel Historian performs better on these structure-derived benchmarks than PRANK,
-which also performs ancestral reconstruction \citep{LoytynojaGoldman2008},
-and is the second-most accurate after ProtPal at estimating indel rates \citep{Westesson2012-zg}.
-Compared to leading protein aligners that do not attempt ancestral reconstruction,
-Indel Historian outperforms CLUSTALW \citep{LarkinEtAl2007} on BaliBase3, but not MUSCLE \citep{Edgar2004b},
-ProbCons \citep{DoEtAl2005}, MAFFT \citep{KatohEtAl2005}, or CLUSTALW on PREFAB (though it is close).
-
-\begin{table}
-  \begin{tabular}{r|rr}
-    & \multicolumn{2}{c}{Average run time per alignment} \\
- & Simulation & BAliBase 3 \\
-    \hline
-Indel Historian & & 55s \\
-PRANK & & 519s \\
-MUSCLE & & 1.9s \\
-\end{tabular}
-\caption{
-    Comparison of runtimes of Indel Historian, PRANK and MUSCLE on the simulation benchmark and the BAliBase 3 benchmark.
-  }
-\end{table}
-
-The results for the programs other than Indel Historian in Table~2 were taken from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
-To make a direct comparison of runtimes, we re-ran the benchmarks for PRANK and MUSCLE on the same machine as the Indel Historian benchmark.
-These runtimes are summarized in Table~3: Indel Historian is an order of magnitude slower than MUSCLE, but an order of magnitude faster than PRANK.
+As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
+To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same machine as the Historian benchmark.
+These runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
+Re-running Prank and Muscle resulted in a slight improvement for Prank
+and a slight deterioration for Muscle compared to the {\tt drive5.com} data,
+presumably due to versioning issues
+(the version of Muscle available for download 3.8.31, whereas the data
+reported are for version 4.0; conversely, a more recent version of Prank is available
+than the one benchmarked on the Muscle website).
+Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.
 
@@
 MCMC samplers may be usefully supplemented by decision-theoretic approaches to summarize a sampling run \citep{HermanEtAl2015}.
-Other potential ways to improve accuracy include position-specific gap penalties, as used by MUSCLE \citep{Edgar2004b} and MAFFT \citep{KatohEtAl2005},
+Other potential ways to improve accuracy include context-dependent gap penalties, as used by Muscle \citep{Edgar2004b}, % and MAFFT \citep{KatohEtAl2005},
 and explicit modeling of tandem duplications \citep{SzalkowskiAnisimova2013}.
@@
 
-% The best-performing method MUSCLE introduces heuristics into its scoring scheme
+% The best-performing method Muscle introduces heuristics into its scoring scheme
 % that improve accuracy.
@@
 
-Indel Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
-It is written in C++11, and compiles on a POSIX system with Clang (v.6.1.0).
-It requires the GNU Scientific Software library (libgsl), the ZLib compression library (libz), and the Boost C++ library.
+Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
+It is written in C++11, compiles on a POSIX system with Clang (v.6.1.0) and requires several free libaries (libgsl, libz and Boost).
+The simulation data reported in this paper are available at \url{github.com/ihh/gp120sim}.
 
@@
 
-Indel Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), % Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort),
+Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), % Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort),
 and the StackOverflow community.

### 2016-10-30 doc/main.tex
Updated

@@
 Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
-Simulation results on fidelity of ancestral sequence reconstruction,
-along with evaluations on the structurally-informed dataset BAliBase 3.0,
-recommend Historian for evolutionary applications.
+Simulation results on fidelity of rate estimation via ancestral reconstruction,
+along with evaluations on the structurally-informed alignment dataset BAliBase 3.0,
+recommend Historian over other alignment tools for evolutionary applications.
 {\bf Availability and Implementation.}
@@
 {\bf Contact.}
-Ian Holmes {\tt ihholmes+indelhistorian@gmail.com}.
+Ian Holmes {\tt ihholmes+historian@gmail.com}.
 {\bf Supplementary Information.}
@@
     ratio (which, for a perfect rate inference, would be equal to 1).
+    Tool versions are as shown in Table~1.
   }
@@
 As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
-To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same machine as the Historian benchmark.
-These runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
+To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same CPU as the Historian benchmark (Intel Xeon 3.20GHz).
+The runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
 Re-running Prank and Muscle resulted in a slight improvement for Prank
@@
 Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
-It is written in C++11, compiles on a POSIX system with Clang (v.6.1.0) and requires several free libaries (libgsl, libz and Boost).
-The simulation data reported in this paper are available at \url{github.com/ihh/gp120sim}.
+Precompiled binaries for Linux or OSX are available.
+The program may also be compiled from the C++11 source
+on a POSIX system with Clang (v.6.1.0) and several free libaries (libgsl, libz and Boost).
+Makefiles to reproduce the analyses reported in this paper
+are included in the repository, and the simulation data
+are available at \url{github.com/ihh/gp120sim}.
 

### 2016-10-30 doc/sim/bin/plotresults.R

@@ -29,11 +29,11 @@ simplot <- function(method,title) {
 		       axis.ticks.y=element_blank()))
 }
 
-ma <- simplot("ma.tree","True alignment")
-prank <- simplot("prank","Prank")
-hist <- simplot("histslow","Historian")
-prob <- simplot("probcons","ProbCons")
-muscle <- simplot("muscle","Muscle")
+ma <- simplot("ma.tree","(a) True alignment")
+hist <- simplot("histslow","(b) Historian")
+prank <- simplot("prank","(c) Prank")
+prob <- simplot("probcons","(d) ProbCons")
+muscle <- simplot("muscle","(e) Muscle")
 
 g <- arrangeGrob(ma,hist,prank,prob,muscle,ncol=1)
 ggsave("results.pdf",g)

### 2016-10-30 doc/main.tex
Abstract
Updated text

@@
 
-\history{Received on XXXXX } %; revised on XXXXX; accepted on XXXXX}
+\history{Received on July 23, 2016; revised on October 30, 2016}
 
-\editor{Associate Editor: XXXXXXX}
+\editor{Associate Editor: Alfonso Valencia}
 
@@
 %are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
-The previous best tool for this purpose was ProtPal,
-but this tool was too slow for practical use.
+The previous best tool for this purpose (according to simulation benchmarks)
+was ProtPal, but this tool was too slow for practical use.
 {\bf Results.}
@@
 while still using the tree of Figure~1.
+Varying rate parameters did not appear
+significantly to affect the ranking of the programs by accuracy.
 
@@
 Historian and Prank are still comparable (with a slight edge to Prank in the
-symmetric tree, vs Historian in the GP120 tree),
+symmetric tree, {\em versus} Historian in the GP120 tree),
 followed by ProbCons, followed by Muscle.
+As with the tree of Figure~1,
+varying rate parameters did not appear to change the ranking of the programs by accuracy.
 

### 2016-10-31 src/recon.cpp
-fast and -faster options

@@ -27,10 +27,12 @@ const regex newick_re (RE_WHITE_OR_EMPTY "\\(" RE_DOT_STAR);  // '('
 const regex json_re (RE_WHITE_OR_EMPTY "\\{" RE_DOT_STAR);    // '{'
 
 const vguard<string> Reconstructor::fastAliasArgs = ReconFastAliasArgs;
+const vguard<string> Reconstructor::fasterAliasArgs = ReconFasterAliasArgs;
 
 Reconstructor::Reconstructor()
   : profileSamples (DefaultProfileSamples),
-    profileNodeLimit (defaultMaxProfileStates()),
+    profileNodeLimit (DefaultProfileMaxStates),
+    maxDPMemoryFraction (DefaultMaxDPMemoryFraction),
     rndSeed (ForwardMatrix::random_engine::default_seed),
     maxDistanceFromGuide (DefaultMaxDistanceFromGuide),
     tokenizeCodons (false),
@@ -68,8 +70,8 @@ Reconstructor::Reconstructor()
     normalizeModel (false)
 { }
 
-int Reconstructor::defaultMaxProfileStates() {
-  return sqrt (DefaultMaxDPMemoryFraction * getMemorySize() / DPMatrix::cellSize());
+int Reconstructor::maxProfileStates() const {
+  return profileNodeLimit ? (int) profileNodeLimit : (int) sqrt (maxDPMemoryFraction * getMemorySize() / DPMatrix::cellSize());
 }
 
 bool Reconstructor::parseAncSeqArgs (deque<string>& argvec) {
@@ -377,6 +379,14 @@ bool Reconstructor::parseProfileArgs (deque<string>& argvec, bool allowReconstru
       argvec.pop_front();
       return true;
 
+    } else if (arg == "-profmaxmem") {
+      Require (argvec.size() > 1, "%s must have an argument", arg.c_str());
+      maxDPMemoryFraction = atof (argvec[1].c_str()) / 100;
+      profileNodeLimit = 0;
+      argvec.pop_front();
+      argvec.pop_front();
+      return true;
+
     } else if (arg == "-nobest") {
       includeBestTraceInProfile = false;
       argvec.pop_front();
@@ -391,6 +401,12 @@ bool Reconstructor::parseProfileArgs (deque<string>& argvec, bool allowReconstru
       argvec.pop_front();
       for (auto fastArgIter = fastAliasArgs.rbegin(); fastArgIter != fastAliasArgs.rend(); ++fastArgIter)
[…]

### 2016-10-31 src/recon.h

@@ -11,7 +11,8 @@
 #include "sampler.h"
 
 #define DefaultProfileSamples 100
-#define DefaultProfilePostProb .01
+#define DefaultProfilePostProb .001
+#define DefaultProfileMaxStates 12000
 #define DefaultMaxDPMemoryFraction .05
 
 #define DefaultMaxDistanceFromGuide 40
@@ -22,7 +23,8 @@
 
 #define AncestralSequencePostProbTag "PP"
 
-#define ReconFastAliasArgs {"-rndspan","-kmatchn","3","-band","10","-profmaxstates","1","-jc","-norefine"}
+#define ReconFastAliasArgs {"-profminpost",".01","-profmaxstates","5000"}
+#define ReconFasterAliasArgs {"-rndspan","-kmatchn","3","-band","10","-profmaxstates","1","-jc","-norefine"}
 
 #define DefaultSimulatorRootSeqLen 100
 
@@ -31,6 +33,7 @@ public:
   typedef AlignColSumProduct::ReconPostProbMap ReconPostProbMap;
 
   static const vguard<string> fastAliasArgs;
+  static const vguard<string> fasterAliasArgs;
   
   string fastaReconFilename, treeFilename, modelFilename, presetModelName;
   list<string> seqFilenames, fastaGuideFilenames, nexusGuideFilenames, stockholmGuideFilenames, nexusReconFilenames, stockholmReconFilenames, countFilenames, simulatorTreeFilenames;
@@ -39,7 +42,7 @@ public:
   size_t profileSamples, profileNodeLimit, maxEMIterations, mcmcSamplesPerSeq;
   int maxDistanceFromGuide, simulatorRootSeqLen, gammaCategories;
   bool tokenizeCodons, guideAlignTryAllPairs, jukesCantorDistanceMatrix, useUPGMA, includeBestTraceInProfile, keepGapsOpen, usePosteriorsForProfile, reconstructRoot, refineReconstruction, predictAncestralSequence, reportAncestralSequenceProbability, accumulateSubstCounts, accumulateIndelCounts, gotPrior, useLaplacePseudocounts, usePosteriorsForDot, useSeparateSubPosteriorsForDot, keepDotGapsOpen, runMCMC, outputTraceMCMC, fixGuideMCMC, fixTreeMCMC, outputLeavesOnly, normalizeModel;
-  double minPostProb, minEMImprovement, minDotPostProb, minDotSubPostProb, gammaShape;
[…]

### 2016-10-31 t/historian.cpp

@@ -109,9 +109,9 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "  -profminpost <P>, -profsamples <N>\n"
     + "                  Specify minimum posterior prob. (P) for retaining DP states\n"
     + "                   in profile (default " + TOSTRING(DefaultProfilePostProb) + "), or sample N traces randomly\n"
-    + "  -profmaxstates <S>\n"
-    + "                  Limit profile to at most S states\n"
-    + "                   (default of " + to_string(Reconstructor::defaultMaxProfileStates()) + " uses at most " + to_string((int) (.5+100*DefaultMaxDPMemoryFraction)) + "% of memory for DP matrix)\n"
+    + "  -profmaxstates <S>, -profmaxmem <M>\n"
+    + "                  Limit profile to at most S states, or to use at most M% of\n"
+    + "                   memory for DP matrix (default is -profmaxstates " + to_string(DefaultProfileMaxStates) + ")\n"
     + "\n"
     + "Following alignment, ancestral sequence reconstruction can be performed.\n"
     + "\n"
@@ -164,8 +164,11 @@ ProgUsage::ProgUsage (int argc, char** argv)
     + "If you are confident the guide alignment & tree should be reasonably obvious,\n"
     + "and just want to get on to reconstruction as quickly as possible:\n"
     + "\n"
-    + "  -fast           Shorthand for the following options:\n"
+    + "  -fast           Run fast. Shorthand for the following:\n"
     + "                   " + join(Reconstructor::fastAliasArgs," ") + "\n"
+    + "\n"
+    + "  -faster         Run even faster. Shorthand for the following:\n"
+    + "                   " + join(Reconstructor::fasterAliasArgs," ") + "\n"
     + "\n"
     + "Model-fitting and event-counting options\n"
     + "~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\n"

### 2016-11-01 doc/main.tex
Updated

@@
 
+The principal parameter determining Historian's execution speed is the number of suboptimal alignments
+it retains while building the progressive ancestral reconstruction. Historian provides a command-line option {\tt -fast} which substantially reduces this number of alignments.
+We here report results for both default- and fast-mode operation.
+
 \end{methods}
@@
 \begin{figure}
-  \includegraphics[width=\columnwidth]{gp120tree.pdf}
+  \includegraphics[width=.8\columnwidth]{gp120tree.pdf}
   \caption{
@@
     For each method evaluated, the distribution of the ratio of inferred to true
-    indel rate is shown, along with the mean and root-mean-squared error for this
-    ratio (which, for a perfect rate inference, would be equal to 1).
+    indel rate is shown; for a perfect rate inference, this ratio would be equal to 1.
+    The mean of the ratio is also shown; this represents the the systematic error,
+    so e.g. using Muscle results in a systematic 15\% underestimate of indels
+    in these simulations.
+    The root-mean-squared value of the ratio is also shown,
+    and the median and interquartile range are annotated above each plot.
     Tool versions are as shown in Table~1.
@@
 (a) using the true evolutionary alignment (supplied to Historian);
-(b) using Historian (in alignment-free mode);
+(b,c) using Historian (alignment-free, default and fast modes);
 and
-(c,d,e) using Prank, ProbCons and Muscle,
+(d,e,f) using Prank, ProbCons and Muscle,
 the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
@@
 Varying rate parameters did not appear
-significantly to affect the ranking of the programs by accuracy.
+significantly to affect the ranking of the programs.
 
@@
 with a slight edge over ProbCons and Muscle.
+Historian's default mode has a slight edge over its fast mode in rate estimation accuracy.
 
@@
 As with the tree of Figure~1,
-varying rate parameters did not appear to change the ranking of the programs by accuracy.
+varying simulation parameters did not appear to change the ranking.
 
@@
     \hline
-Historian v1.0 & 0.82 & 0.50 & \\
-ClustalW & 0.79 & 0.45 & From {\tt drive5.com} \\
-Prank v.100701 & 0.78 & 0.44 \\
-%MAFFT & 0.87 & 0.61 \\
-ProbCons v1.12 & 0.88 & 0.62 & From {\tt drive5.com} \\
-Muscle v3.8.31 & 0.89 & 0.64 & 
+    Historian v1.0.2 & 0.822 & 0.497 & \\
+    (fast mode) & 0.820 & 0.494 & \\
+ClustalW & 0.787 & 0.447 & From {\tt drive5.com} \\
+Prank v.100701 & 0.784 & 0.441 \\
+%MAFFT & 0.873 & 0.607 \\
+ProbCons v1.12 & 0.883 & 0.619 & From {\tt drive5.com} \\
+Muscle v3.8.31 & 0.843 & 0.532 & \\
+Muscle v4.0 & 0.889 & 0.642 &  From {\tt drive5.com}
   \end{tabular}
@@
 or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.
+No significant difference was found between Historian's default and fast modes.
 
@@
   \begin{tabular}{r|c}
-    & Average run time per BAliBase 3.0 alignment \\
+    & Average run time per alignment \\
     \hline
-Historian v1.0 & 55s \\
+Historian v1.0.2 & 233s \\
+ (fast mode) & 55s \\
 Prank v.100701 & 520s \\

### 2016-11-01 doc/sim/bin/plotresults.R

@@ -13,15 +13,25 @@ data.all <- read.csv(file="gp120sim/results.1-100.dat",sep=" ")
 data.sym <- data.all[data.all$tree=="sym",]
 data.tree <- data.all[data.all$tree=="tree",]
 
+quartiles.all <- read.csv(file="gp120sim/quartiles.1-100.dat",sep=" ")
+quartiles.sym <- quartiles.all[quartiles.all$tree=="sym",]
+quartiles.tree <- quartiles.all[quartiles.all$tree=="tree",]
+
 simplot <- function(method,title) {
 	rms <- rms.tree
 	data <- data.tree
+	quartiles <- quartiles.tree
 	err.mean <- rms$insdelmean[rms$method==method]
 	err.rms <- rms$insdelrmse[rms$method==method]
+	err.lq <- quartiles$lower[quartiles$method==method]
+	err.uq <- quartiles$upper[quartiles$method==method]
+	err.median <- quartiles$median[quartiles$method==method]
 	return (ggplot(data[data$method==method,],aes(x=bin,weight=count))
 		+geom_bar()
-		+annotate("text",label=paste0(title,"\nMean error: ",sprintf("%.2f",round(err.mean-1,digits=2)),"\nRMS error: ",sprintf("%.2f",round(err.rms,digits=2))),x=1.6,y=1.75)
-		+xlab("Estimated rate / true rate")
+		+annotate("text",label=paste0(title,"\nMean error: ",sprintf("%.2f",round(err.mean-1,digits=2)),"\nRMS error: ",sprintf("%.2f",round(err.rms,digits=2))),x=1.6,y=1.75,size=3)
+		+geom_segment(aes(y=3,yend=3,x=err.lq,xend=err.uq))
+		+annotate("point",y=3,x=err.median)
+		+xlab("Estimated indel rate / true indel rate")
 		+scale_x_continuous(limits = c(0.25,1.75))
 		+scale_y_continuous(limits = c(0,3))
 		+theme(axis.title.y=element_blank(),
@@ -30,10 +40,13 @@ simplot <- function(method,title) {
 }
 
 ma <- simplot("ma.tree","(a) True alignment")
-hist <- simplot("histslow","(b) Historian")
-prank <- simplot("prank","(c) Prank")
-prob <- simplot("probcons","(d) ProbCons")
-muscle <- simplot("muscle","(e) Muscle")
[…]

### 2016-11-01 doc/sim/bin/results.js

@@ -5,16 +5,30 @@ var fs = require("fs");
 var binsize = process.argv[2],
     binmax = process.argv[3],
     nbins = (binmax/binsize)+1,
-    togetherFlag = eval(process.argv[4]),
+    quartileFlag = eval(process.argv[4]),
     methods = process.argv[5].split(" "),
     dirs = process.argv.slice(6)
 
-function incbin(r,bins) {
+function incbin(r,bins,vals) {
     var bin = Math.round(r/binsize)
     bins[bin] = (bins[bin] || 0) + 1
+    vals.push(r)
 }
 
+process.stdout.write ((quartileFlag
+                       ? ["method","tree","lower","median","upper"]
+                       : ["event","method","rate","tree","multiplier","bin","count"])
+                      .join (" ") + "\n")
+
 var binStart = Array.apply(null,Array(nbins)).map (function(x,n) { return (n*binsize).toPrecision(2) })
+function printrange(method,tree,vals) {
+    var sortedVals = vals.sort (function (a,b) { return a - b })
+    var len = sortedVals.length
+    var lq = sortedVals[Math.floor(len/4)], median = sortedVals[Math.floor(len/2)], uq = sortedVals[Math.floor(3*len/4)]
+    process.stdout.write ([method,tree,lq,median,uq]
+                          .join(" ") + "\n")
+}
+
 function printbins(label,headers,bins) {
     var total = 0
     Object.keys(bins).forEach (function (n) {
@@ -28,40 +42,41 @@ function printbins(label,headers,bins) {
     })
 }
 
-process.stdout.write (["event","method"]
-                      .concat (togetherFlag ? [] : ["rate","tree","multiplier"])
-                      .concat(["bin","count\n"])
-                      .join (" "))
-
 methods.forEach (function(method) {
-    var ibins = {}, dbins = {}
+    var ibins = {}, dbins = {}, vals = {}
     dirs.forEach (function(dir) {
-        var match = /.*rate([\d\.]+)\/([a-z]+)(\d+)/.exec(dir),
-            rate = match[1],
-            tree = match[2],
-            mul = match[3],
-            filenameRegex = new RegExp('\\d+\\.' + method + '\\.json'),
-            headers
-        fs.readdirSync(dir)
-            .filter (function (filename) { return filenameRegex.test(filename) })
-            .map (function (filename) {
-                try {
-                    var json = JSON.parse(fs.readFileSync(dir+"/"+filename));
[…]

### 2016-12-07 doc/main.tex
final paper edit

@@
 leading to a small negative bias in the estimated rate,
-and stochastic noise from a finite sample leads to a spread in the distribution
+and the inherent noisiness of the simulation leads to a spread in the distribution
 of estimated rates from individual alignments.

### 2016-12-11 doc/main.tex
Edited tex source
zipcode
refactored docs

--- 2016-10-30 doc/main.tex
@@
 \usepackage{url}
+\begin{document}
 
-\begin{document}
 \firstpage{1}
 
+\history{Received on July 23, 2016; revised on October 30, 2016}
+\editor{Associate Editor: Alfonso Valencia}
+
 \subtitle{Application Note}
-
 \title[Historian: accurate reconstruction of sequences and rates]{Historian: accurate reconstruction of ancestral sequences and evolutionary rates}
 \author[Ian Holmes]{Ian Holmes$^{\text{\sfb 1}*}$}
-\address{$^{\text{\sf 1}}$Department of Bioengineering, University of California, Berkeley, 94703, USA.}
-
+\address{$^{\text{\sf 1}}$Department of Bioengineering, University of California, Berkeley, 94720, USA.}
 \corresp{$^\ast$To whom correspondence should be addressed.}
 
-\history{Received on July 23, 2016; revised on October 30, 2016}
+\input{paper.tex}
 
-\editor{Associate Editor: Alfonso Valencia}
-
-\abstract{
-{\bf Motivation.}
-Reconstruction of ancestral sequence histories,
-and estimation of parameters like indel rates,
-are improved by using explicit evolutionary models
-and summing over uncertain alignments.
-%The best alignment tools for ancestral reconstruction
-%are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
-The previous best tool for this purpose (according to simulation benchmarks)
-was ProtPal, but this tool was too slow for practical use.
-{\bf Results.}
-Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
-Simulation results on fidelity of rate estimation via ancestral reconstruction,
-along with evaluations on the structurally-informed alignment dataset BAliBase 3.0,
-recommend Historian over other alignment tools for evolutionary applications.
-{\bf Availability and Implementation.}
-Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
-{\bf Contact.}
-Ian Holmes {\tt ihholmes+historian@gmail.com}.
-{\bf Supplementary Information.}
-None.
-}
-
-\maketitle
-
-\section{Introduction}
-
-Multiple alignments are used for several purposes in bioinformatics,
-only one of which is homology-directed protein structure prediction,
-yet this is exactly the application that has tended to dominate alignment benchmarks.
-For evolutionary applications, such as reconstructing trees or ancestral sequences,
-there is evidence that
-different alignment tools (along with different tool-assessment metrics) might be preferable \citep{LoytynojaGoldman2008,Westesson2012-zg}.
-Aligners that are optimized for detecting structural homology may not do so well at recovering information about substitution rates,
-whereas explicit statistical models of the sequence evolution process may do a better job at estimating these parameters.
-Furthermore, the ideal way to estimate evolutionary parameters is not to use a single point estimate of the alignment, but to sum over alignments as a ``nuisance variable''.
-Almost no tools do this, with the exception of MCMC samplers \citep{WestessonBarquistHolmes2012,Redelings2014}.
-
-Empirical studies suggest that, for the purposes of estimating
-molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
-dN/dS ratios \citep{Redelings2014},
-or trees \citep{LoytynojaGoldman2008}---it is advantageous to use a statistical model of evolution
-and to treat alignment rigorously as a ``missing data'' problem.
-%One possible explanation is that, for protein structure prediction,
-%detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
-%(as structurally homologous loops can simply be aligned),
-%whereas evolutionary analyses can make use of this information.
-%
-% The benchmarks commonly used to evaluate aligners use accuracy scores
-% (such as SPS, TCS, AMA, and the Cline shift score)
-% which quantify the number of correctly aligned residues
-% compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
-% Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
-% alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
-% and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
-% hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
-% Aligners have also been tuned for performance by optimizing rate-limiting steps
-% uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
-% 
-In a previous study, we sought to quantify systematic biases introduced into the
-estimation of indel rates, using a simulation benchmark \citep{Westesson2012-zg}.
-The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
-transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
-the least biases of the tools evaluated.
-Unfortunately, the implementation of ProtPal published with that benchmark was too slow
-for practical use.
-
-Here, we present a clean reimplementation of the algorithm underlying ProtPal
-in a new tool, Historian, that can also estimate rates by summing over alignments.
-We report an assessment of the alignment accuracy
-on structurally-derived benchmarks,
-together with a simulation-based experiment to quantify the accuracy of indel rate estimation.
-
-\begin{methods}
-\section{Methods}
-
-Historian combines established algorithms from several sources.
-Like ProtPal, Historian progressively climbs a tree from tips to root,
-building an ancestral sequence profile that includes suboptimal alignments \citep{LeeGrassoSharlow2002,Westesson2012-zg}
-using a time-dependent evolutionary model \citep{RivasEddy2015}.
-The guide tree is found by neighbor-joining % \citep{SaitouNei87}
-on a guide alignment constructed from the all-vs-all pairwise alignment graph,
-or from a sparse random connected subgraph \citep{BradleyEtAl2009}.
-The guide alignment, which can also be supplied by another tool,
-can optionally constrain the progressive reconstruction,
-which is followed by iterative refinement. % \citep{HolmesBruno2001,Edgar2004b}.
-
-Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
-so substitution and indel rates can be estimated directly from sequence data.
-Since the method builds an HMM that captures suboptimal as well as optimal alignments---rather
-like a partial order alignment \citep{LeeGrassoSharlow2002}---the program
-can also estimate rates in an ``alignment-free'' way (i.e. summing over alignments)
-simply by running the Forward-Backward algorithm on this HMM,
-and using the posterior counts to weight the phylo-EM updates.
-
-\end{methods}
-
-\section{Results and Discussion}
-
-\begin{figure}
-  \includegraphics[width=\columnwidth]{gp120tree.pdf}
-  \caption{
-    Tree of selected HIV and SIV GP120 domains, used for the simulation benchmark.
-  }
-\end{figure}
-
-\begin{figure}
-  \includegraphics[width=\columnwidth]{sim/results.pdf}
-  \caption{
-    Results of the simulation benchmark.
-    For each method evaluated, the distribution of the ratio of inferred to true
-    indel rate is shown, along with the mean and root-mean-squared error for this
-    ratio (which, for a perfect rate inference, would be equal to 1).
-    Tool versions are as shown in Table~1.
-  }
-\end{figure}
-
-We first performed a simulation benchmark to assess Historian's ability to reconstruct evolutionary parameters,
-compared to other tools.
-We based the simulation parameters on the evolutionary profile of the HIV/SIV GP120 envelope domain, as follows.
-We started with a tree made from an alignment of ten HIV and SIV GP120 domains (Figure~1).
-%The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
-%The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
-We used the tree to estimate the indel rates for the alignment,
-and simulated 100 different alignments on this tree
-using a sequence length of 500 amino acids
-(roughly the same as the GP120 domain)
-with the expected substitution rate ($\sigma$) set to one substitution per unit of time
-and the simulated insertion and deletion rates ($\lambda,\mu$)
-set to the midpoint of the insertion and deletion rates
-estimated for GP120 (0.028 indels per unit of time).
-For a substitution model, we used the Dayhoff PAM matrix,
-which we judged not to give an unfair advantage either to Historian
-(which by default uses a matrix estimated from PFAM)
-or to Prank (which uses the WAG matrix).
-
-From the simulated sequence data, we estimated indel rates
-(a) using the true evolutionary alignment (supplied to Historian);
-(b) using Historian (in alignment-free mode);
-and
-(c,d,e) using Prank, ProbCons and Muscle,
-the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
-In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
-compared to the simulated indel rates.
-Having performed this experiment
-using the rates estimated for GP120 ($\sigma=1,\lambda=\mu=0.028$),
-we repeated the simulation, varying the rate parameters
-($\sigma=2,\lambda=\mu=0.056;
-\sigma=5,\lambda=\mu=0.14;
-\sigma=2,\lambda=\mu=0.112;
-\sigma=5,\lambda=\mu=0.28;
-\sigma=5,\lambda=\mu=0.005;
-\sigma=5,\lambda=\mu=0.01$)
-while still using the tree of Figure~1.
-Varying rate parameters did not appear
-significantly to affect the ranking of the programs by accuracy.
-
-The results of this experiment are summarized in Figure~2.
-Note that even perfect knowledge of the true evolutionary alignment does not guarantee perfect reconstruction of the underlying indel rate parameters.
-Indel events can overlap and thus be under-counted,
-leading to a small negative bias in the estimated rate,
-and stochastic noise from a finite sample leads to a spread in the distribution
-of estimated rates from individual alignments.
-Historian and Prank (which both explicitly aim to provide ancestral reconstructions)
-are the most accurate methods for rate estimation,
-with a slight edge over ProbCons and Muscle.
-
-As well as varying the rate parameters,
-we also repeated the benchmark on a symmetric 8-taxon binary tree.
-This simulation yielded similar patterns:
-Historian and Prank are still comparable (with a slight edge to Prank in the
-symmetric tree, {\em versus} Historian in the GP120 tree),
-followed by ProbCons, followed by Muscle.
-As with the tree of Figure~1,
-varying rate parameters did not appear to change the ranking of the programs by accuracy.
-
-\begin{table}
-  \begin{tabular}{r|ccc}
-    & Mean SPS & Mean TCS & Notes \\
-    \hline
-Historian v1.0 & 0.82 & 0.50 & \\
-ClustalW & 0.79 & 0.45 & From {\tt drive5.com} \\
-Prank v.100701 & 0.78 & 0.44 \\
-%MAFFT & 0.87 & 0.61 \\
-ProbCons v1.12 & 0.88 & 0.62 & From {\tt drive5.com} \\
-Muscle v3.8.31 & 0.89 & 0.64 & 
-  \end{tabular}
-  \caption{
-    Comparison of Historian to other alignment programs using the BAliBase 3.0 benchmark
-    with SPS and TCS alignment quality scores \citep{ThompsonEtAl2005}.
-%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
-%    Data for ClustalW, Prank (+F option, ClustalW tree), MAFFT (v6.603), ProbCons (v1.12) and Muscle (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
-  }
-\end{table}
-
-Table~2 summarizes an evaluation of Historian
-on the BAliBase 3.0 benchmark,
-compiled using 3D structural alignments.
-%
-In general, Historian performs better on these structure-derived benchmarks than Prank,
-which also performs ancestral reconstruction \citep{LoytynojaGoldman2008}.
-Compared to leading protein aligners that do not attempt ancestral reconstruction,
-Historian outperforms ClustalW \citep{LarkinEtAl2007}, but not Muscle \citep{Edgar2004b}
-or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.
-
-\begin{table}
-  \begin{tabular}{r|c}
-    & Average run time per BAliBase 3.0 alignment \\
-    \hline
-Historian v1.0 & 55s \\
-Prank v.100701 & 520s \\
-Muscle v3.8.31 & 1.9s \\
-\end{tabular}
-\caption{
-    Comparison of runtimes of Historian, Prank and Muscle on the BAliBase 3.0 benchmark.
-  }
-\end{table}
-
-As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
-To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same CPU as the Historian benchmark (Intel Xeon 3.20GHz).
-The runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
-Re-running Prank and Muscle resulted in a slight improvement for Prank
-and a slight deterioration for Muscle compared to the {\tt drive5.com} data,
-presumably due to versioning issues
-(the version of Muscle available for download 3.8.31, whereas the data
-reported are for version 4.0; conversely, a more recent version of Prank is available
-than the one benchmarked on the Muscle website).
-Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.
-
-We did not benchmark MCMC approaches
-such as BaliPhy \citep{Redelings2014}, StatAlign \citep{NovakEtAl2008} or HandAlign \citep{WestessonBarquistHolmes2012}.
-These are expected to be more accurate, but generally take much longer.
-MCMC samplers may be usefully supplemented by decision-theoretic approaches to summarize a sampling run \citep{HermanEtAl2015}.
-Other potential ways to improve accuracy include context-dependent gap penalties, as used by Muscle \citep{Edgar2004b}, % and MAFFT \citep{KatohEtAl2005},
-and explicit modeling of tandem duplications \citep{SzalkowskiAnisimova2013}.
-Incorporating additional data such as structural annotations may further improve alignments \citep{HermanEtAl2014}.
-
-% The best-performing method Muscle introduces heuristics into its scoring scheme
-% that improve accuracy.
-% 
-% It is harder to incorporate these sorts of modifications into a rigorous phylogenetic model
-% primarily because they violate the assumption of independence
-% between indel and substitution processes
-% that most such models rely on
-% 
-% Comparison to MCMC tools
-% BaliPhy \citep{RedelingsSuchard2005,RedelingsSuchard2007,Redelings2014},
-% StatAlign \citep{NovakEtAl2008,HermanEtAl2014},
-% HandAlign \citep{WestessonBarquistHolmes2012}
-% 
-% Decision theory may also be useful to produce consensus alignments summarizing an MCMC run \citep{HermanEtAl2015}
-
-\section{Availability}
-
-Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
-Precompiled binaries for Linux or OSX are available.
-The program may also be compiled from the C++11 source
-on a POSIX system with Clang (v.6.1.0) and several free libaries (libgsl, libz and Boost).
-Makefiles to reproduce the analyses reported in this paper
-are included in the repository, and the simulation data
-are available at \url{github.com/ihh/gp120sim}.
-
-\section*{Acknowledgements}
-
-Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), % Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort),
-and the StackOverflow community.
-
-\section*{Funding}
-
-This work has been supported by NHGRI grant R01-HG004483.
-
-\bibliographystyle{natbib}
-%\bibliographystyle{bioinformatics}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%
-%\bibliographystyle{plain}
-%
-%\bibliography{Document}
-
-\bibliography{references}
 \end{document}

### 2016-12-11 doc/paper.tex
refactored docs

\abstract{
{\bf Motivation.}
Reconstruction of ancestral sequence histories,
and estimation of parameters like indel rates,
are improved by using explicit evolutionary models
and summing over uncertain alignments.
%The best alignment tools for ancestral reconstruction
%are not necessarily those that agree most closely with benchmark datasets based on 3D structure alignments.
The previous best tool for this purpose (according to simulation benchmarks)
was ProtPal, but this tool was too slow for practical use.
{\bf Results.}
Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
Simulation results on fidelity of rate estimation via ancestral reconstruction,
along with evaluations on the structurally-informed alignment dataset BAliBase 3.0,
recommend Historian over other alignment tools for evolutionary applications.
{\bf Availability and Implementation.}
Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
{\bf Contact.}
Ian Holmes {\tt ihholmes+historian@gmail.com}.
{\bf Supplementary Information.}
None.
}

\maketitle

\section{Introduction}

Multiple alignments are used for several purposes in bioinformatics,
only one of which is homology-directed protein structure prediction,
yet this is exactly the application that has tended to dominate alignment benchmarks.
For evolutionary applications, such as reconstructing trees or ancestral sequences,
there is evidence that
different alignment tools (along with different tool-assessment metrics) might be preferable \citep{LoytynojaGoldman2008,Westesson2012-zg}.
Aligners that are optimized for detecting structural homology may not do so well at recovering information about substitution rates,
whereas explicit statistical models of the sequence evolution process may do a better job at estimating these parameters.
Furthermore, the ideal way to estimate evolutionary parameters is not to use a single point estimate of the alignment, but to sum over alignments as a ``nuisance variable''.
Almost no tools do this, with the exception of MCMC samplers \citep{WestessonBarquistHolmes2012,Redelings2014}.

Empirical studies suggest that, for the purposes of estimating
molecular evolutionary parameters---such as indel rates \citep{Westesson2012-zg},
dN/dS ratios \citep{Redelings2014},
or trees \citep{LoytynojaGoldman2008}---it is advantageous to use a statistical model of evolution
and to treat alignment rigorously as a ``missing data'' problem.
%One possible explanation is that, for protein structure prediction,
%detailed reconstruction of indel histories in fast-evolving regions (such as loops) is unnecessary
%(as structurally homologous loops can simply be aligned),
%whereas evolutionary analyses can make use of this information.
%
% The benchmarks commonly used to evaluate aligners use accuracy scores
% (such as SPS, TCS, AMA, and the Cline shift score)
% which quantify the number of correctly aligned residues
% compared to a dataset of structurally-informed reference alignments \citep{ThompsonEtAl2005}.
% Techniques that score highly in these tests include using Bayesian decision theory to maximize the expected
% alignment accuracy \citep{NotredameEtAl2000,DoEtAl2005,SchwartzPachter2007,BradleyEtAl2009}
% and fine-tuning the alignment scoring function for contextual details, such as the reduced propensity for indels in
% hydrophobic protein regions \citep{Edgar2004b,KatohEtAl2005,LarkinEtAl2007}.
% Aligners have also been tuned for performance by optimizing rate-limiting steps
% uch as all-versus-all pairwise sequence comparison \citep{Edgar2004b,BradleyEtAl2009}.
% 
In a previous study, we sought to quantify systematic biases introduced into the
estimation of indel rates, using a simulation benchmark \citep{Westesson2012-zg}.
The ProtPal program \citep{Westesson2012-zg}, which models indel evolution using 
transducers---finite-state machines which can be multiplied together like substitution matrices \citep{BouchardCote2013}---introduced
the least biases of the tools evaluated.
Unfortunately, the implementation of ProtPal published with that benchmark was too slow
for practical use.

Here, we present a clean reimplementation of the algorithm underlying ProtPal
in a new tool, Historian, that can also estimate rates by summing over alignments.
We report an assessment of the alignment accuracy
on structurally-derived benchmarks,
together with a simulation-based experiment to quantify the accuracy of indel rate estimation.

%\begin{methods}
\section{Methods}

Historian combines established algorithms from several sources.
Like ProtPal, Historian progressively climbs a tree from tips to root,
building an ancestral sequence profile that includes suboptimal alignments \citep{LeeGrassoSharlow2002,Westesson2012-zg}
using a time-dependent evolutionary model \citep{RivasEddy2015}.
The guide tree is found by neighbor-joining % \citep{SaitouNei87}
on a guide alignment constructed from the all-vs-all pairwise alignment graph,
or from a sparse random connected subgraph \citep{BradleyEtAl2009}.
The guide alignment, which can also be supplied by another tool,
can optionally constrain the progressive reconstruction,
which is followed by iterative refinement. % \citep{HolmesBruno2001,Edgar2004b}.

Historian also implements the phylogenetic EM algorithm for continuous-time Markov chains \citep{HolmesRubin2002},
so substitution and indel rates can be estimated directly from sequence data.
Since the method builds an HMM that captures suboptimal as well as optimal alignments---rather
like a partial order alignment \citep{LeeGrassoSharlow2002}---the program
can also estimate rates in an ``alignment-free'' way (i.e. summing over alignments)
simply by running the Forward-Backward algorithm on this HMM,
and using the posterior counts to weight the phylo-EM updates.

The principal parameter determining Historian's execution speed is the number of suboptimal alignments
it retains while building the progressive ancestral reconstruction. Historian provides a command-line option {\tt -fast} which substantially reduces this number of alignments.
We here report results for both default- and fast-mode operation.

%\end{methods}

\section{Results and Discussion}

\begin{figure}
  \includegraphics[width=.8\columnwidth]{figure1.pdf}
  \caption{
    Tree of selected HIV and SIV GP120 domains, used for the simulation benchmark.
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure2.pdf}
  \caption{
    Results of the simulation benchmark.
    For each method evaluated, the distribution of the ratio of inferred to true
    indel rate is shown; for a perfect rate inference, this ratio would be equal to 1.
    The mean of the ratio is also shown; this represents the the systematic error,
    so e.g. using Muscle results in a systematic 15\% underestimate of indels
    in these simulations.
    The root-mean-squared value of the ratio is also shown,
    and the median and interquartile range are annotated above each plot.
    Tool versions are as shown in Table~1.
  }
\end{figure}

We first performed a simulation benchmark to assess Historian's ability to reconstruct evolutionary parameters,
compared to other tools.
We based the simulation parameters on the evolutionary profile of the HIV/SIV GP120 envelope domain, as follows.
We started with a tree made from an alignment of ten HIV and SIV GP120 domains (Figure~1).
%The resulting tree has three branches that are over 0.1 substitutions/site in length (the relevant branch lengths are 0.10, 0.19, and 0.41).
%The remaining branch lengths were all under 0.1 subs/site; the median branch length was 0.046 subs/site.
We used the tree to estimate the indel rates for the alignment,
and simulated 100 different alignments on this tree
using a sequence length of 500 amino acids
(roughly the same as the GP120 domain)
with the expected substitution rate ($\sigma$) set to one substitution per unit of time
and the simulated insertion and deletion rates ($\lambda,\mu$)
set to the midpoint of the insertion and deletion rates
estimated for GP120 (0.028 indels per unit of time).
For a substitution model, we used the Dayhoff PAM matrix,
which we judged not to give an unfair advantage either to Historian
(which by default uses a matrix estimated from PFAM)
or to Prank (which uses the WAG matrix).

From the simulated sequence data, we estimated indel rates
(a) using the true evolutionary alignment (supplied to Historian);
(b,c) using Historian (alignment-free, default and fast modes);
and
(d,e,f) using Prank, ProbCons and Muscle,
the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
compared to the simulated indel rates.
Having performed this experiment
using the rates estimated for GP120 ($\sigma=1,\lambda=\mu=0.028$),
we repeated the simulation, varying the rate parameters
($\sigma=2,\lambda=\mu=0.056;
\sigma=5,\lambda=\mu=0.14;
\sigma=2,\lambda=\mu=0.112;
\sigma=5,\lambda=\mu=0.28;
\sigma=5,\lambda=\mu=0.005;
\sigma=5,\lambda=\mu=0.01$)
while still using the tree of Figure~1.
Varying rate parameters did not appear
significantly to affect the ranking of the programs.

The results of this experiment are summarized in Figure~2.
Note that even perfect knowledge of the true evolutionary alignment does not guarantee perfect reconstruction of the underlying indel rate parameters.
Indel events can overlap and thus be under-counted,
leading to a small negative bias in the estimated rate,
and the inherent noisiness of the simulation leads to a spread in the distribution
of estimated rates from individual alignments.
Historian and Prank (which both explicitly aim to provide ancestral reconstructions)
are the most accurate methods for rate estimation,
with a slight edge over ProbCons and Muscle.
Historian's default mode has a slight edge over its fast mode in rate estimation accuracy.

As well as varying the rate parameters,
we also repeated the benchmark on a symmetric 8-taxon binary tree.
This simulation yielded similar patterns:
Historian and Prank are still comparable (with a slight edge to Prank in the
symmetric tree, {\em versus} Historian in the GP120 tree),
followed by ProbCons, followed by Muscle.
As with the tree of Figure~1,
varying simulation parameters did not appear to change the ranking.

\begin{table}
  \begin{tabular}{r|ccc}
    & Mean SPS & Mean TCS & Notes \\
    \hline
    Historian v1.0.2 & 0.822 & 0.497 & \\
    (fast mode) & 0.820 & 0.494 & \\
ClustalW & 0.787 & 0.447 & From {\tt drive5.com} \\
Prank v.100701 & 0.784 & 0.441 \\
%MAFFT & 0.873 & 0.607 \\
ProbCons v1.12 & 0.883 & 0.619 & From {\tt drive5.com} \\
Muscle v3.8.31 & 0.843 & 0.532 & \\
Muscle v4.0 & 0.889 & 0.642 &  From {\tt drive5.com}
  \end{tabular}
  \caption{
    Comparison of Historian to other alignment programs using the BAliBase 3.0 benchmark
    with SPS and TCS alignment quality scores \citep{ThompsonEtAl2005}.
%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
%    Data for ClustalW, Prank (+F option, ClustalW tree), MAFFT (v6.603), ProbCons (v1.12) and Muscle (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
  }
\end{table}

Table~2 summarizes an evaluation of Historian
on the BAliBase 3.0 benchmark,
compiled using 3D structural alignments.
%
In general, Historian performs better on these structure-derived benchmarks than Prank,
which also performs ancestral reconstruction \citep{LoytynojaGoldman2008}.
Compared to leading protein aligners that do not attempt ancestral reconstruction,
Historian outperforms ClustalW \citep{LarkinEtAl2007}, but not Muscle \citep{Edgar2004b}
or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.
No significant difference was found between Historian's default and fast modes.

\begin{table}
  \begin{tabular}{r|c}
    & Average run time per alignment \\
    \hline
Historian v1.0.2 & 233s \\
 (fast mode) & 55s \\
Prank v.100701 & 520s \\
Muscle v3.8.31 & 1.9s \\
\end{tabular}
\caption{
    Comparison of runtimes of Historian, Prank and Muscle on the BAliBase 3.0 benchmark.
  }
\end{table}

As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same CPU as the Historian benchmark (Intel Xeon 3.20GHz).
The runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
Re-running Prank and Muscle resulted in a slight improvement for Prank
and a slight deterioration for Muscle compared to the {\tt drive5.com} data,
presumably due to versioning issues
(the version of Muscle available for download 3.8.31, whereas the data
reported are for version 4.0; conversely, a more recent version of Prank is available
than the one benchmarked on the Muscle website).
Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.

We did not benchmark MCMC approaches
such as BaliPhy \citep{Redelings2014}, StatAlign \citep{NovakEtAl2008} or HandAlign \citep{WestessonBarquistHolmes2012}.
These are expected to be more accurate, but generally take much longer.
MCMC samplers may be usefully supplemented by decision-theoretic approaches to summarize a sampling run \citep{HermanEtAl2015}.
Other potential ways to improve accuracy include context-dependent gap penalties, as used by Muscle \citep{Edgar2004b}, % and MAFFT \citep{KatohEtAl2005},
and explicit modeling of tandem duplications \citep{SzalkowskiAnisimova2013}.
Incorporating additional data such as structural annotations may further improve alignments \citep{HermanEtAl2014}.

% The best-performing method Muscle introduces heuristics into its scoring scheme
% that improve accuracy.
% 
% It is harder to incorporate these sorts of modifications into a rigorous phylogenetic model
% primarily because they violate the assumption of independence
% between indel and substitution processes
% that most such models rely on
% 
% Comparison to MCMC tools
% BaliPhy \citep{RedelingsSuchard2005,RedelingsSuchard2007,Redelings2014},
% StatAlign \citep{NovakEtAl2008,HermanEtAl2014},
% HandAlign \citep{WestessonBarquistHolmes2012}
% 
% Decision theory may also be useful to produce consensus alignments summarizing an MCMC run \citep{HermanEtAl2015}

\section{Availability}

Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
Precompiled binaries for Linux or OSX are available.
The program may also be compiled from the C++11 source
on a POSIX system with Clang (v.6.1.0) and several free libaries (libgsl, libz and Boost).
Makefiles to reproduce the analyses reported in this paper
are included in the repository, and the simulation data
are available at \url{github.com/ihh/gp120sim}.

\section*{Acknowledgements}

Historian includes code from Ivan Vashchaev (gason), Heng Li (kseq), % Rafael Baptista (stacktrace), David Robert Nadeau (memsize), Roger Pate (string escaping), Konrad Rudolph (index sort),
and the StackOverflow community.

\section*{Funding}

This work has been supported by NHGRI grant R01-HG004483.

\bibliographystyle{natbib}
%\bibliographystyle{bioinformatics}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\bibliography{references}

### 2016-12-11 doc/paper.tex
refactored docs

@@
-\abstract{
-{\bf Motivation.}
+\structabs{
 Reconstruction of ancestral sequence histories,
@@
 was ProtPal, but this tool was too slow for practical use.
-{\bf Results.}
+}{
 Historian combines an efficient reimplementation of the ProtPal algorithm with performance-improving heuristics from other alignment tools.
@@
 recommend Historian over other alignment tools for evolutionary applications.
-{\bf Availability and Implementation.}
+}{
 Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
-{\bf Contact.}
+}{
 Ian Holmes {\tt ihholmes+historian@gmail.com}.
-{\bf Supplementary Information.}
+}{
 None.

### 2016-12-13 doc/paper.tex
refactored docs
Updated URL, license

@@
 }{
-Historian is available at \url{https://github.com/ihh/indelhistorian} under the Creative Commons Attribution 3.0 US license.
+Historian is available at \url{https://github.com/evoldoers/historian} under the BSD3 license.
 }{
@@
 }
-
-\maketitle
 
@@
 
-Historian is available at \url{github.com/ihh/indelhistorian} under the CC BY 3.0 US license.
+Historian is available at \url{https://github.com/evoldoers/historian} under the BSD3 license.
 Precompiled binaries for Linux or OSX are available.

### 2016-12-15 perl/gamma-sites.pl
renamed indelhistorian -> historian

@@ -28,7 +28,7 @@ my ($shape, $scale, $bins) = (1, 1, 2);
 my ($pretty, $verbose) = (0, 0);
 
 my $usage = "";
-$usage .= "$progname -- add discretized-gamma prior to indelhistorian rate matrix\n";
+$usage .= "$progname -- add discretized-gamma prior to historian rate matrix\n";
 $usage .= "\n";
 $usage .= "Usage: $progname [opts] params.json\n";
 $usage .= "\n";

### 2016-12-28 doc/paper.tex
App note edits

@@
 
-As noted in Table~2, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
+As noted in Table~1, some of the results were taken from the Muscle website, {\tt drive5.com/bench/} \citep{Edgar2010}.
 To make a direct comparison of runtimes, we re-ran the benchmarks for Prank and Muscle on the same CPU as the Historian benchmark (Intel Xeon 3.20GHz).
-The runtimes are summarized in Table~3: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
+The runtimes are summarized in Table~2: Historian is an order of magnitude slower than Muscle, but an order of magnitude faster than Prank.
 Re-running Prank and Muscle resulted in a slight improvement for Prank
@@
 than the one benchmarked on the Muscle website).
-Table~2 reflects our results, where available, and the {\tt drive5.com} results in other cases.
+Table~1 reflects our results, where available, and the {\tt drive5.com} results in other cases.
 
